---
title: "Basic Diversity Example"
author: "Bob O'Hara"
format:
  html
bibliography: references.bib
execute: 
  cache: true
---

# Introduction

This is an implementation of the diversity statistics, to demonstrate their calculation without any trait or environmental variables. We will use data where we can estimate both occurrence and abundance (conditional on occurrence).

# The Community Model


For species $i$ on site $j$ we denote occurrence as $z_{ij} \in \{0,1\}$ (where 1 is presence), and abundance as $\lambda_{ij} > 0$. We will link both of these to the general theory, where we assume the occurrence/abundance/whatever is $k_ij$ for species $j$ on site $i$.

Fo occurrence we assume presence is a Bernoulli trial, then $z_{ij} \sim \text{Bern}(p_{ij})$.  with probability $\text{logit}(\eta_{ij}) = \log\frac{p_{ij}}{1-p_{ij}}$. This is equivalent to a threshold model, i.e. saying that we have a continuous random variable, $\eta_{ij}^{(z)}$, and $z_{ij} = 1$ if $\eta_{ij}^{(z)}>0$. We will assume that $\eta_{ij}^{(z)}$ follows a logistic distribution, which is equivalent to using a logit link in a GLM. Then 

$$
\eta_{ij}^{(z)} = \mu^{(z)} + \alpha^{(z)}_i + \beta^{(z)}_j + \gamma^{(z)}_j x_i + \delta^{(z)}_{ij}.
$$

And $\delta^{(z)}_{ij}$ follows a standard logistic distribution, which means it has a mean of 0 and variance $\pi^2/3$. We can calculate statistics on $\eta_{ij}^{(z)}$: note that $z_{ij}|\eta_{ij}^{(z)}$ is deterministic: the usual Bernoulli stochasticity is in $\delta^{(z)}_{ij}$. 

For abundance we assume a log-linear model:

$$
\log(\lambda_{ij}|z_{ij}=1) = \eta_{ij}^{(\lambda)} = \mu^{(\lambda)} + \alpha^{(\lambda)}_i + \beta^{(\lambda)}_j + \gamma^{(\lambda)}_j x_i + \delta^{(\lambda)}_{ij}
$$

Clearly $\lambda_{ij}=0$ if $z_{ij}=0$.

For both models we assume the effects are random effects, i.e. with a variance to be estimated. But to make the presentation easier to follow, we use the posterior estimates of the parameters to calculate variances from. 


The models are fitted with JAGS, so it is easier to estimate uncertainty. Detailsa re in the fold-out below.



# An Example



::: {.callout-note collapse=true}

## Observation Model

The data are counts of each species, and we model both presence/absence and counts. We have multiple replicate observations per site, so we assume the counts, $n_{kj}$ on replicate $k$ for species $j$, follow a zero inflated Poisson lognormal distribution with mean $\lambda_{ij}$ (i.e. constant across replicates within a site). If the detection probability is constant within a species (but differing between species) this will change the $\beta^{(\lambda)}_j$ term. We have a Gaussian overdispersion term for excess variation within a replicate that represents sampling variation.

## Downloading and formatting the data

The data are abundances (counts) on each location on a transect. There are 3 islands, each with 3 transects, each with 3 stations. Each station was sampled in 2 seasons, and has 3 replicates.

The steps are:
1. import the data

```{r}
#| echo: false
Env <- read.csv("data/waddenX.csv")
Counts <- read.csv("data/waddenY.csv")
Species <- names(Counts)[!names(Counts)%in%c("X", "measurementID")]

# Extract information about data
Obs <- strsplit(Counts$measurementID, "_")

Counts$Island <- unlist(lapply(Obs, function(x) x[[1]]))
Counts$Transect <- unlist(lapply(Obs, function(x) x[[2]]))
Counts$Station <- unlist(lapply(Obs, function(x) x[[3]]))
Counts$Season <- unlist(lapply(Obs, function(x) x[[4]]))
Counts$Replicate <- unlist(lapply(Obs, function(x) x[[5]]))

Counts$Site <- factor(paste(Counts$Island, Counts$Transect, Counts$Station, Counts$Season, sep="_"))

Present <- apply(Counts[,Species], 2, function(Spp, Site) 
  tapply(Spp, list(Site), function(a) c(NA,1)[1+any(a>0)]), 
                 Site = Counts$Site)

```

2. Format the data for JAGS.

```{r}
#| echo: false
DataToJAGS <- list(
  Abund = as.matrix(Counts[,Species]), 
  Site = as.numeric(Counts$Site),
  Present = Present, 
  temperature = as.vector(scale(tapply(Env$temperature, list(Counts$Site), mean))),
  # Island <- Counts$Island, # Might use these later
  # Transect <- Counts$Transect, 
  # Station <- Counts$Station, 
  # Season <- Counts$Season

  NSp = length(Species),
  NSite = nlevels(Counts$Site),
  NRep = nrow(Counts)
)

# Function to simulate initial values
GetInits <- function(data, corr=FALSE) {
  res <- list(
    mu.be.lam = rnorm(1, 0, 0.1),
    mu.be.z = rnorm(1, 0, 0.1),
    mu.al.lam = rnorm(1, 0, 0.1),
    mu.al.z = rnorm(1, 0, 0.1),
    mu.lam = rnorm(1, 0, 0.1),
    mu.z = rnorm(1, 0, 0.1),
    
    sd.be.lam = rgamma(1, 100, 500),
    sd.be.z = rgamma(1, 100, 500),
    sd.al.lam = rgamma(1, 100, 500),
    sdOD = rgamma(1, 100, 500),
    sd.al.z = rgamma(1, 100, 500))
  if(exists("temperature", data)) {
    res$mu.gam.lam = rnorm(1, 0, 1)
    res$sd.gam.lam = rgamma(1, 100, 500)
    res$mu.gam.z = rnorm(1, 0, 0.5)
    res$sd.gam.z = rgamma(1, 100, 500)
  }
  if(corr) {
    res$rho = rnorm(data$NSp, 0, 0.1)
    res$VarSiteOD = matrix(rgamma(2*data$NSp, 100, 100), ncol=2)
  }
  res
}

```

3. Write the model in JAGS code

```{r}
#| echo: false
cat("
model {
  for(sp in 1:NSp) {
    for(rep in 1:NRep) {
      Abund[rep, sp] ~ dpois(lambda[rep, sp])
      lambda[rep, sp] <- mu[rep, sp]*Present[Site[rep], sp]
      OD[rep, sp] ~ dnorm(0, tauOD)
      log(mu[rep, sp]) <- logmu[Site[rep], sp] + OD[rep, sp]
    }
    for(site in 1:NSite) {
      Present[site, sp] ~ dbern(p[site, sp])
      SiteOD[site, sp] ~ dmnorm(0, tauSiteOD)
      logmu[site, sp] <- mu.lam + al.lamST[site] + be.lamST[sp] + SiteOD[site, sp]
      logit(p[site, sp]) <- mu.z + al.zST[site] + be.zST[sp]
    }
    be.zST[sp] ~ dnorm(mu.be.z, tau.be.z)
    be.lamST[sp] ~ dnorm(mu.be.lam, tau.be.lam)

    be.lam[sp] <- be.lamST[sp] - mean(be.lamST[])
    be.z[sp] <- be.zST[sp] - mean(be.zST[])
  }
  for(site in 1:NSite) {
    al.lamST[site] ~ dnorm(mu.al.lam, tau.al.lam)
    al.zST[site] ~ dnorm(mu.al.z, tau.al.z)
    al.lam[site] <- al.lamST[site] - mean(al.lamST[])
    al.z[site] <- al.zST[site] - mean(al.zST[])
    SR[site] <- sum(Present[site, ])
  }
    
  mu.be.lam ~ dnorm(0, 1)
  mu.be.z ~ dnorm(0, 1)
  mu.al.lam ~ dnorm(0, 1)
  mu.al.z ~ dnorm(0, 1)
  mu.lam ~ dnorm(0, 1)
  mu.z ~ dnorm(0, 1)

  Mean.lam <- mu.lam + mean(al.lamST[]) + mean(be.lamST[])
  Mean.z <- mu.z + mean(al.zST[]) + mean(be.zST[])

  tau.be.lam <- pow(sd.be.lam, -2); sd.be.lam ~ dexp(1)
  tau.be.z <- pow(sd.be.z, -2); sd.be.z ~ dexp(1)
  tau.al.lam <- pow(sd.al.lam, -2); sd.al.lam ~ dexp(1)
  tau.al.z <- pow(sd.al.z, -2); sd.al.z ~ dexp(1)

  tauOD <- pow(sdOD, -2); sdOD ~ dexp(1)
  tauSiteOD <- pow(sdSiteOD, -2); sdSiteOD ~ dexp(1)

# Finite population variances
  StdDev.be.lam <- sd(be.lamST[])
  StdDev.be.z <- sd(be.zST[])
  StdDev.al.lam  <- sd(al.lamST[])
  StdDev.al.z <- sd(al.zST[])

  StdDevSiteOD <- sd(SiteOD[, ])
  StdDevDelta.lam <- sd(SiteOD[, ]) # interaction
  StdDevDelta.z <- sqrt(3.289868) # interaction
  StdDevOD <- sd(OD[, ]) # observation level OD
  StdDevTotal.z <- sqrt(StdDev.al.z^2 + StdDev.be.z^2 + StdDevDelta.z^2)
  StdDevTotal.lam <- sqrt(StdDev.al.lam^2 + StdDev.be.lam^2 + StdDevDelta.lam^2)

}", file="BeatingBetaCode.jag")

```



4. Run the JAGS code

```{r}
#| warning: false
#| message: false
#| output: false
#| echo: false
library(runjags)
library(coda)

ToMonitor.z <- c("Mean.z", "SR", "sd.al.z", "sd.be.z", "al.z")
ToMonitor.lam <- c("Mean.lam", "sd.al.lam", "sd.be.lam", "sdSiteOD", "sdOD", "StdDevDelta")
ToMonitor.gam <- c("mu.gam.lam", "sd.gam.lam", "sd.gam.z", "mu.gam.z", 
                   "gamma.lam", "gamma.z", "Mean.gam.lam", "Mean.gam.z")
ToMonitor.means <- c("logmu", "Present", "p") # if we want everything
ToMonitor.DivStats <- c("Unevenness", "UnevennessOcc")
ToMonitor.Unevenness <- c("be.lam", "SiteOD", "Present")

ToMonitor <- unique(c(ToMonitor.z, ToMonitor.lam, ToMonitor.gam, ToMonitor.means,
                      ToMonitor.DivStats))
# Add to this as we go on
ToMonitor <- unique(c(ToMonitor.means))

Inits <- replicate(2, GetInits(DataToJAGS), simplify = FALSE)

DivModel <- run.jags(model = "BeatingBetaCode.jag", monitor=ToMonitor, data=DataToJAGS,
                  inits = Inits, method="parallel", silent.jags=TRUE,
#         n.chains = 2, adapt = 1e2, burnin = 0, sample = 150)
#         n.chains = 4, adapt = 1e3, burnin = 0, sample = 5e3)
         n.chains = 6, adapt = 1e3, burnin = 0, sample = 1e3)
# Summ <- summary(DivModel, vars= c(ToMonitor.z, ToMonitor.lam, ToMonitor.gam))
# Summ <- summary(DivModel, vars= ToMonitor)

# Combine MCMC into one data fame, so it's easier to work with below
DivPars <- combine.mcmc(DivModel)

# this is a function to calculate the summaries we want
CalcSumm <- function(mc) {
  require(coda)
  require(MCMCglmm)
  Summ <- summary(mc)
  Summ$HPD <- HPDinterval(mc, prob=0.95)
  Summ$PMode <- posterior.mode(mc)
  Summ
}


```

We get some output, that we should probably check for convergence.

:::

# Diversity Output

All of our diversity metrics are based on decomposing the distributions into:

- Occurrence
- Abundance

And looking at the means and variances of the parameters associated with these.

## Within-Site Diversity

## Within-Site Means: $\beta$-mean diversity

The overall means are probably rather boring, but we can look at the site means. For abundance we only average over the species that are present: for those we do not observe, we estimate presence.

```{r}
Sites <- 1:DataToJAGS$NSite
Species <- 1:DataToJAGS$NSp

# Function to Calculate mean occupancy or abundance for each site
CalcSiteMean <- function(site, mc, Occ=TRUE) {
# Extract relevant statistics from MCMC draws
  mu <- mc[,grep(paste0("^logmu\\[", site, ","), colnames(mc))]
  inc <- mc[,grep(paste0("^Present\\[", site, ","), colnames(mc))]
  p <- mc[,grep(paste0("^p\\[", site, ","), colnames(mc))]
  
  if(Occ) {
    res <- apply(log(1/(1/p - 1)), 1, mean)
  } else {
    res <- sapply(1:nrow(mc), function(i, mu, pres) {
      mean(mu[1,pres[i,]==1]) # mean for when presence=1
    }, mu=mu, pres=inc)
  }
  res
  
}

# Function to calculate species richness
CalcSR <- function(site, mc) {
  inc <- mc[,grep(paste0("^Present\\[", site, ","), colnames(mc))]
  res <- apply(inc, 1, sum)
  res
}

RsiteOcc <- coda::as.mcmc(sapply(Sites, CalcSiteMean, Occ=TRUE, mc=DivPars))
RsiteAbund <- coda::as.mcmc(sapply(Sites, CalcSiteMean, Occ=FALSE, mc=DivPars))
SpRichness <- coda::as.mcmc(sapply(Sites, CalcSR, mc=DivPars))


SummRsiteOcc <- CalcSumm(RsiteOcc)
SummRsiteAbund <- CalcSumm(RsiteAbund)
SummSpRichness <- CalcSumm(SpRichness)
```

The estimates are plotted below, with the same site on the same row.

```{r}
#| fig-height: 8
#| fig-cap: Posterior modes and 95% HPDIs for unevenness of communities on sites

PlotPars <- function(Summ, At="Orig",...) {
  if(length(At)==length(Summ$PMode)) {
    At.y <- At
  } else {
    At.y <- switch(At, 
                   Orig = 1:length(Summ$PMode), 
                   rank = rank(Summ$PMode, ties="random")
    )
  }
  plot(Summ$PMode, At.y, xlim=range(Summ$HPD), ...)
  segments(Summ$HPD[,"lower"], At.y, Summ$HPD[,"upper"], At.y)
  
}

At.mean <- rank(SummRsiteOcc$PMode, ties="random")
par(mfrow=c(1,3), mar=c(2,2,4,1))
PlotPars(SummRsiteOcc, At=At.mean, main="Mean Occupancy")
PlotPars(SummSpRichness, At=At.mean, main="Species Richness")
PlotPars(SummRsiteAbund, At=At.mean, main="Mean Abundance")
```

Not surprisingly, the posterior modes for species richness and mean occupancy are strongly correlated ($\rho$: `r round(cor(SummRsiteOcc$PMode, SummSpRichness$PMode), 2)`). But the occurrence and abundance are prett much uncorrelated ($\rho$: `r round(cor(SummRsiteOcc$PMode, SummRsiteAbund$PMode), 2)`).


### Unevenness: α-var-diversity

Our theory says that unevenness is $\text{Var}_j(k_{i,j})$, i.e. the variance in the species' effects on each site. For abundance this is related to measures such as Shannon's and Simpson's diversity: the more evenly distributed species' abundances are, the higher the abundance (an uneven distribution would mean one or a few species with high abundances, i.e. dominating the community). Our variance is a measure of unevenness: when it is zero, all species have the same abundance, so there is maximal diversity.

In practice we calculate our abundance unevenness only using the species estimated to be present on the site.

```{r}
Sites <- 1:DataToJAGS$NSite
Species <- 1:DataToJAGS$NSp

CalcUnevenness <- function(site, mc, Occ=TRUE) {
  mu <- mc[,grep(paste0("^logmu\\[", site, ","), colnames(mc))]
  inc <- mc[,grep(paste0("^Present\\[", site, ","), colnames(mc))]
  p <- mc[,grep(paste0("^p\\[", site, ","), colnames(mc))]
  
  if(Occ) {
    res <- apply(log(1/(1/p - 1)), 1, var)
  } else {
    res <- sapply(1:nrow(mc), function(i, mu, pres) {
      var(mu[1,pres[i,]==1])
    }, mu=mu, pres=inc)
  }
  res
  
}

UnevennessOcc <- sapply(Sites[1], CalcUnevenness, mc=DivPars)
UnevennessAbund <- coda::as.mcmc(sapply(Sites, CalcUnevenness, Occ=FALSE, mc=DivPars))

library(MCMCglmm)

SummUn <- summary(UnevennessAbund)
SummUn$HPD <- HPDinterval(UnevennessAbund, prob=0.95)
SummUn$PMode <- posterior.mode(UnevennessAbund)

```

For occurrence there is only one unevenness

```{r}

plot(density(UnevennessOcc), main="Occurrence Unevenness", xlab="Unevenness")

```


```{r}
#| fig-height: 8
#| fig-cap: Posterior modes and 95% HPDIs for unevenness of communities on sites

At.y <- rank(SummUn$PMode)
plot(SummUn$PMode, At.y, xlim=range(SummUn$HPD), xlab="Unevenness")
segments(SummUn$HPD[,"lower"], At.y, SummUn$HPD[,"upper"], At.y)

```

## Species Variation: $\beta$-mean diversity

We can also look at species. 

```{r}
Sites <- 1:DataToJAGS$NSite
Species <- 1:DataToJAGS$NSp

# Function to Calculate mean occupancy or abundance for each site
CalcSpMean <- function(sp, mc, Occ=TRUE) {
# Extract relevant statistics from MCMC draws
  mu <- mc[,grep(paste0("^logmu\\[.*,", sp, "]"), colnames(mc))]
  inc <- mc[,grep(paste0("^Present\\[.*,", sp, "]"), colnames(mc))]
  p <- mc[,grep(paste0("^p\\[.*,", sp, "]"), colnames(mc))]
  
  if(Occ) {
    res <- apply(log(1/(1/p - 1)), 1, mean)
  } else {
    res <- sapply(1:nrow(mc), function(i, mu, pres) {
      mean(mu[1,pres[i,]==1]) # mean for when presence=1
    }, mu=mu, pres=inc)
  }
  res
  
}

# Function to calculate species range (i.e. how many sites the species is present on)
CalcSpRange <- function(sp, mc) {
  inc <- mc[,grep(paste0("^Present\\[.*,", sp, "]"), colnames(mc))]
  res <- apply(inc, 1, sum)
  res
}

SpMeanOcc <- coda::as.mcmc(sapply(Species, CalcSpMean, Occ=TRUE, mc=DivPars))
SpMeanAbund <- coda::as.mcmc(sapply(Species, CalcSpMean, Occ=FALSE, mc=DivPars))
SpRange <- coda::as.mcmc(sapply(Species, CalcSpRange, mc=DivPars))

SummSpMeanOcc <- CalcSumm(SpMeanOcc)
SummSpMeanAbund <- CalcSumm(SpMeanAbund)
SummSpRange <- CalcSumm(SpRange)

```


```{r}

At.sp <- rank(SummSpMeanOcc$PMode, ties="random")
par(mfrow=c(1,3), mar=c(2,2,4,1))
PlotPars(SummSpMeanOcc, At=At.sp, main="Mean Occupancy")
PlotPars(SummSpRange, At=At.sp, main="Range Size")
PlotPars(SummSpMeanAbund, At=At.sp, main="Mean Abundance")

```

The posterior modes for range size and mean occupancy are strongly correlated ($\rho$: `r round(cor(SummSpMeanOcc$PMode, SummSpRange$PMode), 2)`). The occurrence and abundance have a low correlation ($\rho$: `r round(cor(SummSpMeanOcc$PMode, SummSpMeanAbund$PMode), 2)`).

### Specificity: β-var-diversity

In practice we calculate our abundance unevenness only using the species estimated to be present on the site.

```{r}
Sites <- 1:DataToJAGS$NSite
Species <- 1:DataToJAGS$NSp

CalcSpecificity <- function(sp, mc, Occ=TRUE) {
  mu <- mc[,grep(paste0("^logmu\\[.*,", sp, "]"), colnames(mc))]
  inc <- mc[,grep(paste0("^Present\\[.*,", sp, "]"), colnames(mc))]
  p <- mc[,grep(paste0("^p\\[.*,", sp, "]"), colnames(mc))]
  
  if(Occ) {
    res <- apply(log(p/(1-p)), 1, var)
  } else {
    res <- sapply(1:nrow(mc), function(i, mu, inc) {
      mus <- mu[i,inc[i,]==1]
      ifelse(length(mus)>1, var(mus), 0)
    }, mu=mu, inc=inc)
  }
  res
  
}

SpecificityOcc <- sapply(Species[1], CalcSpecificity, mc=DivPars)
SpecificityAbund <- coda::as.mcmc(sapply(Species, CalcSpecificity, Occ=FALSE, mc=DivPars))


thing <- CalcSpecificity(Species[58], Occ=FALSE, mc=DivPars)

SummSpecAbund <- CalcSumm(SpecificityAbund)


```

For occurrence there is only one specificity

```{r}

plot(density(SpecificityOcc), main="Occurrence Specificity", xlab="Specificity")

```

We can plot the specificity for abundance. There is considerable uncertainty. The species are sorted by their posterior means.

```{r}
#| fig-height: 8
#| fig-cap: Posterior modes and 95% HPDIs for specificty of species


At.spec <- rank(SummSpecAbund$statistics[,"Mean"], ties="random")
PlotPars(SummSpecAbund, At=At.sp, main="Mean Occupancy")


```

If we plot the densities, we see that most of them are bimodal, with a mode at or around 0. We get 0 if the species was present on one site.

```{r}
#| fig-height: 15
#| fig-cap: "Posterior densities for species specificity in abundance"
par(mfrow=c(15,4), mar=c(2,2,1,1), oma=c(2,2,0,0))
sapply(Species, function(wh, mc) plot(density(mc[,wh]), main=""),
       mc=SpecificityAbund)
mtext("Specificity", 1, outer=TRUE)
mtext("Density", 2, outer=TRUE)

```

# Beta Diversity

