---
title: "Basic Diversity Example"
author: "Bob O'Hara"
format:
  pdf
bibliography: references.bib
execute: 
  cache: true
---

# Introduction

This is an implementation of the diversity statistics, to demonstrate their calculation without any trait or environmental variables. We will use data whre we can estimate both occurrence and abundance (conditional on occurrence)

# The Community Model

We For species $i$ on site $j$ we denote occurrence as $z_{ij} \in \{0,1\}$, and abundance as $\lambda_{ij}$.

For the general theory we assume the incidence/abundance/whatever is $k_ij$ for species $j$ on site $i$. Here we have $k_{ij}$ as one of $\{ z_{ij}, \lambda_{ij}\}$, where $z_{ij}$ is presence/absence, and $\lambda_{ij}$ is the actual abundance on site $i$. 

We assume presence is a Bernoulli trial, so if $z_{ij} \in \{0,1\}$, where 1 means presence, then $z_{ij} \sim \text{Bern}(p_{ij})$, with probability $\text{logit}(p_{ij}) = \log\frac{p_{ij}}{1-p_{ij}}$. This is equivalent to a threshold model, i.e. saying that we have a continuous random variable, $\eta_{ij}^{(z)}$, and $z_{ij} = 1$ if $\eta_{ij}^{(z)}>0$. We will assume that $\eta_{ij}^{(z)}$ follows a logistic distribution, which is equivalent to using a logit link in a GLM. Then 

$$
\eta_{ij}^{(z)} = \mu^{(z)} + \alpha^{(z)}_i + \beta^{(z)}_j + \gamma^{(z)}_j x_i + \delta^{(z)}_{ij}.
$$

And $\delta^{(z)}_{ij}$ follows a standard logistic distribution, which means it has a mean of 0 and variance $\pi^2/3$. We can calculate statistics on $\eta_{ij}^{(z)}$: note that $z_{ij}|\eta_{ij}^{(z)}$ is deterministic: the usual Bernoulli stochasticity is in $\delta^{(z)}_{ij}$. 

For abundance we assume a log-linear model:

$$
\log(\lambda_{ij}|z_{ij}=1) = \eta_{ij}^{(\lambda)} = \mu^{(\lambda)} + \alpha^{(\lambda)}_i + \beta^{(\lambda)}_j + \gamma^{(\lambda)}_j x_i + \delta^{(\lambda)}_{ij}
$$

Clearly $\lambda_{ij}=0$ if $z_{ij}=0$.

For both models we assume the effects are random effects, i.e. with a variance to be estimated, and we can assume that $E(\alpha^{(z)}_i)=E(\beta^{(z)}_j)=E(\delta^{(z)}_{ij})=E(\alpha^{(\lambda)}_i)=E(\beta^{(\lambda)}_j)=E(\delta^{(\lambda)}_{ij})=0$, i.e. the intercept effects are all mean centred. $\gamma^{(z)}_j$ and $\gamma^{(\lambda)}_j$ have means to be estimated (i.e. the average effect of $x_i$ over all species).

## Variances

Going from the above, if we write the linear predictor as $k_{ij}$, for either $z$ or $\lambda$, and ignore the superscripts, we get

$$
k_{ij} = \mu + \lambda_i + \eta_j  + \delta_{ij}
$$

From which (following Chrisoph's work) we see that the total diversity (gamma diversity) is 

$$
  \text{Var}_{ij}(k_{ij}) = \sigma_{\lambda}^2 + \sigma_{\eta}^2 + \sigma_{\delta}^2
$$

Where we write $\text{Var}_{ij}(k_{ij})$ to be explicit that we are calculating the variance over both sites ($i$) and species ($j$). 

<!---
## Finite Population Variances and Infinite Population Variances

In the statistical modelling there are two types of variance we will come across. If, for example, we assume $\alpha_i \sim N(0,\sigma_\alpha^2)$, we have $\sigma_\alpha^2$, the infinite population variance, but we can also calculate $\text{Var}(\alpha_i) = \frac{1}{m}\sum \alpha_i^2$, the finite population variance. The difference is whether we only consider the observations or the (infinite) population that these observations are assumed to have been drawn from, These, and particularly their sampling distributions, may not be the same. 

The difference between these is that the finite population variance is the variance of what was measured, i.e. the variance of the samples from the population. The infinite population variance is the variance over all possible samples. If the samples are random (and the model is correct), the two should be equivalent. Which one to use will depend on the questions being asked, i.e. if they are about the specific locations being sampled (in which case the finite population variance is appropriate), or if the questions are about the whole region, including unsampled sites (when the infinite population variance should be used).

One area where this becomes an issue is with compositional alpha diversity, where we can estimate $\text{Var}(\delta_{ij}|i)$, even if we assume the infinite population variance is constant across sites.

At the moment the population and super population estimates are mixed up, but they will be separated later (I only thought about this once I had written most of the code, so it will need some additions).
--->

# Diversity Measures

## Means 

Incidence




## Alpha Diversity


**Species Richness**: 

**Unevenness**: 

## Beta Diversity

Beta diversity is variation between sites. As with alpha diversity, we can first split it into species presence and composition. We can go further and follow @Baselga2010 by decomposing the species presence component into changes in richness ("nestedness") and changes in which species are present (which we call "replacement").

### Species Identity

Here we call total variation in species identity "turnover", and it is made up of nestedness & replacement. Replacement is, roughly, the change in species species composition if the species richness was the same. 


### Within Species

Within species diversity is akin to alpha diversity, but now it is about the incidence and abundance of one species across several sites. These are the possible statistics, for which we need some interpetations:

- Range size: $R_j = \sum_i z_{ij}$, i.e. the number of sites the species is observed on (note that this is not the same as geographic range). We could also use $\mu^{(z)}+\beta^{(z)}_j$.
- Site specificity: $\text{Var}_{i}(\eta^{(z)}_{ij}|j) = \sigma^{2}_{(z)\alpha} + \pi/3$. This is the variance in presence across sites, so is a measure of habitat specificity: if it is high, the species is always found on some sites, and never on others. If it is low, it's random where we find it.
- mean local abundance: $E_i(\eta^{(\lambda)}_{ij}|j) = \mu^{(\lambda)}+\beta^{(\lambda)}_j$
- Variance in abundance: $\text{Var}_{i}(\eta^{(\lambda)}_{ij}|j) = \sigma^{2}_{(\lambda)\alpha} + \sigma^{2}_{(\lambda)\gamma}$. This is also related to site specificity.

Site richness is related to having a restricted distribution, mean abundance is related to scarcity. We can link this to the typology of rarity from @CrisfieldRarity, as range size (which modifies geographic range _sensu_  @rabinowitz1981Rarity), local abundance and occupancy (which replaces habitat specificity in the scheme of  @rabinowitz1981Rarity). These can be identified with our within species statistics.


# An Example

::: {.callout-note collapse=true}

## Observation Model

The data are counts of each species, and we model both presence/absence and counts. We have multiple replicate observations per site, so we assume the counts, $n_{kj}$ on replicate $k$ for species $j$, follow a zero inflated Poisson lognormal distribution with mean $\lambda_{ij}$ (i.e. constant across replicates within a site). If the detection probability is constant within a species (but differing between species) this will change the $\beta^{(\lambda)}_j$ term. We have a Gaussian overdispersion term for excess variation within a replicate that represents sampling variation.

## Downloading and formatting the data

The data are abundances (counts) on each location on a transect. There are 3 islands, each with 3 transects, each with 3 stations. Each station was sampled in 2 seasons, and has 3 replicates.

The steps are:
1. import the data

```{r}
#| echo: false
Env <- read.csv("data/waddenX.csv")
Counts <- read.csv("data/waddenY.csv")
Species <- names(Counts)[!names(Counts)%in%c("X", "measurementID")]

# Extract information about data
Obs <- strsplit(Counts$measurementID, "_")

Counts$Island <- unlist(lapply(Obs, function(x) x[[1]]))
Counts$Transect <- unlist(lapply(Obs, function(x) x[[2]]))
Counts$Station <- unlist(lapply(Obs, function(x) x[[3]]))
Counts$Season <- unlist(lapply(Obs, function(x) x[[4]]))
Counts$Replicate <- unlist(lapply(Obs, function(x) x[[5]]))

Counts$Site <- factor(paste(Counts$Island, Counts$Transect, Counts$Station, Counts$Season, sep="_"))

Present <- apply(Counts[,Species], 2, function(Spp, Site) 
  tapply(Spp, list(Site), function(a) c(NA,1)[1+any(a>0)]), 
                 Site = Counts$Site)

```

2. Format the data for JAGS.

```{r}
#| echo: false
DataToJAGS <- list(
  Abund = as.matrix(Counts[,Species]), 
  Site = as.numeric(Counts$Site),
  Present = Present, 
  temperature = as.vector(scale(tapply(Env$temperature, list(Counts$Site), mean))),
  # Island <- Counts$Island, # Might use these later
  # Transect <- Counts$Transect, 
  # Station <- Counts$Station, 
  # Season <- Counts$Season

  NSp = length(Species),
  NSite = nlevels(Counts$Site),
  NRep = nrow(Counts)
)

# Function to simulate initial values
GetInits <- function(data, corr=FALSE) {
  res <- list(
    mu.be.lam = rnorm(1, 0, 0.1),
    mu.be.z = rnorm(1, 0, 0.1),
    mu.al.lam = rnorm(1, 0, 0.1),
    mu.al.z = rnorm(1, 0, 0.1),
    mu.lam = rnorm(1, 0, 0.1),
    mu.z = rnorm(1, 0, 0.1),
    
    sd.be.lam = rgamma(1, 100, 500),
    sd.be.z = rgamma(1, 100, 500),
    sd.al.lam = rgamma(1, 100, 500),
    sdOD = rgamma(1, 100, 500),
    sd.al.z = rgamma(1, 100, 500))
  if(exists("temperature", data)) {
    res$mu.gam.lam = rnorm(1, 0, 1)
    res$sd.gam.lam = rgamma(1, 100, 500)
    res$mu.gam.z = rnorm(1, 0, 0.5)
    res$sd.gam.z = rgamma(1, 100, 500)
  }
  if(corr) {
    res$rho = rnorm(data$NSp, 0, 0.1)
    res$VarSiteOD = matrix(rgamma(2*data$NSp, 100, 100), ncol=2)
  }
  res
}

```

3. Write the model in JAGS code

```{r}
#| echo: false
cat("
model {
  for(sp in 1:NSp) {
    for(rep in 1:NRep) {
      Abund[rep, sp] ~ dpois(lambda[rep, sp])
      lambda[rep, sp] <- mu[rep, sp]*Present[Site[rep], sp]
      OD[rep, sp] ~ dnorm(0, tauOD)
      log(mu[rep, sp]) <- logmu[Site[rep], sp] + OD[rep, sp]
    }
    for(site in 1:NSite) {
      Present[site, sp] ~ dbern(p[site, sp])
      SiteOD[site, sp] ~ dmnorm(0, tauSiteOD)
      logmu[site, sp] <- mu.lam + al.lamST[site] + be.lamST[sp] + SiteOD[site, sp] + 
          gamma.lam[sp]*(temperature[site]-mean(temperature[]))
      logit(p[site, sp]) <- mu.z + al.zST[site] + be.zST[sp] + 
          gamma.z[sp]*(temperature[site]-mean(temperature[]))
    }
    be.lamST[sp] ~ dnorm(mu.be.lam, tau.be.lam)
    be.zST[sp] ~ dnorm(mu.be.z, tau.be.z)
    gamma.lam[sp] ~ dnorm(mu.gam.lam, tau.gam.lam)
    gamma.z[sp] ~ dnorm(mu.gam.z, tau.gam.z)

    be.lam[sp] <- be.lamST[sp] - mean(be.lamST[])
    be.z[sp] <- be.zST[sp] - mean(be.zST[])
    # These are deviations
    gamma.lamDev[sp] <- gamma.lam[sp] - mean(gamma.lam[])
    gamma.zDev[sp] <- gamma.z[sp] - mean(gamma.z[])
  }
  for(site in 1:NSite) {
    al.lamST[site] ~ dnorm(mu.al.lam, tau.al.lam)
    al.zST[site] ~ dnorm(mu.al.z, tau.al.z)
    al.lam[site] <- al.lamST[site] - mean(al.lamST[])
    al.z[site] <- al.zST[site] - mean(al.zST[])
    SR[site] <- sum(Present[site, ])
    temperature[site] ~ dnorm(0, 1)
    StdDevDelta[site] <- sd(SiteOD[site, ])
    
# Means 
    Rz[site] <- al.z[site] + Mean.z
    Rlam[site] <- al.lam[site] + Mean.lam + inprod(be.lam,Present[site, ])
# Site Variances

  }  
  
  for(sp in 1:NSp) {
  # Species Means
    Oz[sp] <- be.z[sp] + Mean.z
    Olam[sp] <- be.lam[sp] + Mean.lam + inprod(al.lam,Present[, sp])
  # Species Variances
  
  }

  mu.be.lam ~ dnorm(0, 1)
  mu.be.z ~ dnorm(0, 1)
  mu.al.lam ~ dnorm(0, 1)
  mu.al.z ~ dnorm(0, 1)
  mu.lam ~ dnorm(0, 1)
  mu.z ~ dnorm(0, 1)
  mu.gam.lam ~ dnorm(0, 1)
  mu.gam.z ~ dnorm(0, 1)

  Mean.lam <- mu.lam + mean(al.lamST[]) + mean(be.lamST[])
  Mean.z <- mu.z + mean(al.zST[]) + mean(be.zST[])

  Mean.gam.lam <- mu.gam.lam + mean(gamma.lam[])
  Mean.gam.z <- mu.z
  
  tau.be.lam <- pow(sd.be.lam, -2); sd.be.lam ~ dexp(1)
  tau.be.z <- pow(sd.be.z, -2); sd.be.z ~ dexp(1)
  tau.al.lam <- pow(sd.al.lam, -2); sd.al.lam ~ dexp(1)
  tau.al.z <- pow(sd.al.z, -2); sd.al.z ~ dexp(1)

  tau.gam.lam <- pow(sd.gam.lam, -2); sd.gam.lam ~ dexp(1)
  tau.gam.z <- pow(sd.gam.z, -2); sd.gam.z ~ dexp(1)
  tauOD <- pow(sdOD, -2); sdOD ~ dexp(1)
  tauSiteOD <- pow(sdSiteOD, -2); sdSiteOD ~ dexp(1)

# Finite population variances
  StdDev.be.lam <- sd(be.lamST[])
  StdDev.be.z <- sd(be.zST[])
  StdDev.al.lam  <- sd(al.lamST[])
  StdDev.al.z <- sd(al.zST[])

  StdDev.gam.lam <- sd(gamma.lam[])
  StdDev.gam.z <- sd(gamma.z[])
  StdDevSiteOD <- sd(SiteOD[, ])
#  Note: temperature is scaled
  StdDevDelta.lam <- sqrt(sd(SiteOD[, ])^2 +  (mu.gam.lam^2 + sd(gamma.lam)^2)) # interaction
  StdDevDelta.z <- sqrt(3.289868 +  (mu.gam.z^2 + sd(gamma.z)^2)) # interaction
  StdDevOD <- sd(OD[, ]) # observation level OD
  StdDevTotal.z <- sqrt(StdDev.al.z^2 + StdDev.be.z^2 + StdDevDelta.z^2)
  StdDevTotal.lam <- sqrt(StdDev.al.lam^2 + StdDev.be.lam^2 + StdDevDelta.lam^2)

}", file="BeatingBetaCode.jag")

```



4. Run the JAGS code

```{r}
#| warning: false
#| message: false
#| output: false
#| echo: false
library(runjags)
library(coda)

ToMonitor.z <- c("Mean.z", "SR", "sd.al.z", "sd.be.z", "al.z")
ToMonitor.lam <- c("Mean.lam", "sd.al.lam", "sd.be.lam", "sdSiteOD", "sdOD", "StdDevDelta")
ToMonitor.gam <- c("mu.gam.lam", "sd.gam.lam", "sd.gam.z", "mu.gam.z", 
                   "gamma.lam", "gamma.z", "Mean.gam.lam", "Mean.gam.z")
ToMonitor.means <- c("logmu", "Present") # if we want everything
ToMonitor.DivStats <- c("Mean.z", "Mean.lam", "Rz", "Rlam", "Oz", "Olam",
                        "sd.al.z", "sd.be.z", "sd.al.lam", "sd.be.lam",
                        "StdDevDelta.z", "StdDevDelta.lam",
                        "StdDevTotal.z", "StdDevTotal.lam")
ToMonitor <- unique(c(ToMonitor.z, ToMonitor.lam, ToMonitor.gam, ToMonitor.means))
Inits <- replicate(2, GetInits(DataToJAGS), simplify = FALSE)

DivModel <- run.jags(model = "BeatingBetaCode.jag", monitor=ToMonitor, data=DataToJAGS,
                  inits = Inits, method="parallel", silent.jags=TRUE,
#         n.chains = 2, adapt = 1e2, burnin = 0, sample = 150)
#         n.chains = 4, adapt = 1e3, burnin = 0, sample = 5e3)
         n.chains = 6, adapt = 1e3, burnin = 0, sample = 1e3)
Summ <- summary(DivModel, vars= c(ToMonitor.z, ToMonitor.lam, ToMonitor.gam))

# Combine MCMC into one data fame, so it's easier to work with below
DivPars <- combine.mcmc(DivModel)

```

We get some output. in the fold below are some trace plots so we can see if it has converged & if it is mixing well. The caution is well warranted.

:::

# Diversity Output

All of our diversity metrics are based on decomposing the distributions into:

- Incidence
- Composition

And looking at the means and variances of the parameters associated with these.



## Incidence

Alpha diversity is a within-site diversity. We split it into species richness and composition, and look at unevenness as a summary of composition.

### Means and what they mean

(1) **We have $\mu$, the grand mean**

```{r}
#| echo: false

plot(DivModel, vars="Mean.gam.z", plot.type = "histogram")

```

(2) **Species Richness**

We calculate species richness on site $i$, $S_i$, as $\sum_j z_{ij}$

```{r}
#| echo: false
GetSpRichness <- function(mc) {
  SR <- mc[,grep("SR", colnames(mc))]
  Rz <- mc[,grep("al.z", colnames(mc))] + mc[,grep("Mean.z", colnames(mc))]
  as.mcmc(SR, Rz)
}
SR <- GetSpRichness(mc=DivPars)
# summary(SR)

```

In @fig-SR we plot the estimated species richness for each site for the data, with colours for islands and open/closed circles for summer/spring respectively.

```{r}
#| fig-height: 10
#| fig-cap: "Posterior estimates of species richness. Filled circles: spring, open circles: summer."
#| label: fig-SR
#| echo: false
SRRows <- grep("SR", rownames(Summ))
CodeIsland <- as.numeric(factor(gsub("_.*", "", rownames(Present))))
CodeSeason <- as.numeric(factor(gsub(".*_", "", rownames(Present))))

par(mar=c(4.1,4.1,1,1))
plot(Summ[SRRows,"Mean"], SRRows, col=CodeIsland, 
     pch=c(1,16)[CodeSeason], cex=2, 
     xlim=c(0, max(Summ[SRRows,"Upper95"])),
     xlab="Species Richness", ylab="Site")
segments(Summ[SRRows,"Lower95"], SRRows, Summ[SRRows,"Upper95"], SRRows, 
         col=CodeIsland)
legend(1, 18, c("Nor", "Sp", "Wang"), col=1:3, lwd=2)

```

(3) **Average Abundance**



```{r}
#| echo: false
# GetSpRichness <- function(mc) {
#   SR <- mc[,grep("SR", colnames(mc))]
#   Rz <- DivPars[,grep("al.lam", colnames(mc))] + mc[,grep("Mean.lam", colnames(mc))]
#   as.mcmc(SR, Rz)
# }
Rz <- as.mcmc(DivPars[,grep("al.lam", colnames(DivPars))] + 
                DivPars[,grep("Mean.lam", colnames(DivPars))])

# SR <- GetSpRichness(mc=DivPars)
# summary(SR)

```

In @fig-AveSiteAbund we plot the estimated average site abundance

```{r}
#| fig-height: 10
#| fig-cap: "Posterior estimates of species richness. Filled circles: spring, open circles: summer."
#| label: fig-AveSiteAbund
#| echo: false
SRRows <- grep("SR", rownames(Summ))
CodeIsland <- as.numeric(factor(gsub("_.*", "", rownames(Present))))
CodeSeason <- as.numeric(factor(gsub(".*_", "", rownames(Present))))

par(mar=c(4.1,4.1,1,1))
plot(Summ[SRRows,"Mean"], SRRows, col=CodeIsland, 
     pch=c(1,16)[CodeSeason], cex=2, 
     xlim=c(0, max(Summ[SRRows,"Upper95"])),
     xlab="Species Richness", ylab="Site")
segments(Summ[SRRows,"Lower95"], SRRows, Summ[SRRows,"Upper95"], SRRows, 
         col=CodeIsland)
legend(1, 18, c("Nor", "Sp", "Wang"), col=1:3, lwd=2)

```

(4) Prevalence (?) Habitat specificity?

This is the average occupancy per species

```{r}
#| echo: false
GetSiteMeans <- function(mc) {
  Oz <- mc[,grep("be.z", colnames(mc))] + mc[,grep("Mean.z", colnames(mc))]
  Olam <- mc[,grep("be.lam", colnames(mc))] + mc[,grep("Mean.lam", colnames(mc))]
  as.mcmc(cbind(Oz, Olam))
}
SR <- GetSpRichness(mc=DivPars)
# summary(SR)

```



### Unevenness

Our estimate of (site specific) unevenness is 

$$
\text{Var}_{j}(\eta_{ij}|i) = \sigma_{\beta}^2 + \sigma_{\gamma}^2 x_i + \sigma_{\delta,i}^2
$$

We plot it for our sites in @fig-UnevennessData. The unevennesses look very high, apparently because of rare species. We can look at the distribution of posterior means for a couple of sites (grey is the total distribution of present species, red is for species that are observed to be present) in @fig-SiteDistr.

```{r}
#| fig-height: 10
#| label: fig-UnevennessData
#| fig-cap: "Posterior estimates of unevenness for all sites. Filled circles: spring, open circles: summer."
#| warning: false
#| echo: false
# If site=FALSE this is the same for every site
CalcUnevenness <- function(mc, XX, site=FALSE, prop=FALSE) {
  if(any(is.na(XX))) {
    XX[is.na(XX)] <- 0
    message("Some covariates are missing: these have been set to 0")
  }
  resX <- mc[,grep("sd.gam.lam", colnames(mc))]%o%XX^2
  if(site) {
    eta <- mc[,grep("logmu", colnames(mc))]
    Z <- mc[,grep("Present", colnames(mc))]
    Site.eta <- gsub(",.*", "", gsub(".*\\[", "", colnames(eta)))
    Species.eta <- gsub(".*,", "", gsub("\\]", "", colnames(eta)))
    
    res <- sapply(unique(Site.eta), function(site, eta, pres, siteInd) {
      ind <- pres[,siteInd==site]==1
      etaS <- eta[,siteInd==site]
      StdDev <- sapply(1:nrow(etaS), function(row, eta, ind) sd(eta[row,ind[row,]]), 
                       eta=etaS, ind=ind)
      StdDev
    }, eta=eta, pres=Z, siteInd=Site.eta)
  } else {
    res <- mc[,grep("sd.be.lam", colnames(mc))]^2 + resX + 
      mc[,grep("sdSiteOD", colnames(mc))]
  }
  if(prop) { # Does this make sense?
    prop <- resX/res
    res <- cbind(res, prop)
  }
  as.mcmc(res)
}
Unevenness <- CalcUnevenness(mc=DivPars, XX=DataToJAGS$temperature, prop=FALSE, site=TRUE)
SummSD <- summary(Unevenness)
logmuSDRows <- 1:nrow(SummSD$statistics)

par(mar=c(4.1,4.1,1,1))
plot(SummSD$statistics[,"Mean"], logmuSDRows, col=CodeIsland, 
     pch=c(1,16)[CodeSeason], cex=2, 
     xlim=c(min(SummSD$quantiles[,"2.5%"]), max(SummSD$quantiles[,"97.5%"])),
     xlab="Unevenness", ylab="Site")
segments(SummSD$quantiles[,"2.5%"], logmuSDRows, 
         SummSD$quantiles[,"97.5%"], logmuSDRows, 
         col=CodeIsland)
legend(1.9, 13, c("Nor", "Sp", "Wang"), col=1:3, lwd=2)

```

Most species that are modelled to be present are really rare. 

```{r}
#| fig-height: 8
#| echo: false
#| label: fig-SiteDistr
#| fig-cap: "Estimates of abundance distributions for four sites: estimates for species which were observed are in red, and unobserved species (i.e. those with abundance 0 on these sites, but where z=1) in grey."
# site,species
# mcmc <- DivModel$mcmc[[1]]
GetlogMuSite <- function(mcmc, site=1) {
  UseColsPres <- colnames(mcmc)[grep(paste0("Present\\[", site, ","), colnames(mcmc))]
  Present.mcmc <- mcmc[,UseColsPres]

  UseColsAbund <- colnames(mcmc)[grep(paste0("logmu\\[", site, ","), colnames(mcmc))]
  logmu.mcmc <- mcmc[,UseColsAbund]+(1e4*(1-Present.mcmc))
  logmu.mcmc
}

HistMeanMus <- function(mcmcs, site=1, Pres=NULL, ...) {
  lmus <- do.call(rbind, lapply(mcmcs, GetlogMuSite, site=site))

  HistAll <- hist(apply(lmus, 2, function(x) mean(x[x<1e3])), plot=FALSE)
  
  if(!is.null(Pres)) {
    HistPres <- hist(apply(lmus[,!is.na(Pres[site,])], 2, function(x) mean(x[x<1e3])), 
                     breaks=HistAll$breaks, plot=FALSE)
  }
  plot(HistAll, ...)
  if(!is.null(Pres)) plot(HistPres, add = TRUE, col=2)
}

par(mfrow=c(2,2), oma=c(2,2,0,0), mar=c(2.6,2.1,4.1,2))
HistMeanMus(DivModel$mcmc, site=2, Pres = DataToJAGS$Present, main="Site 2", xlab="", ylab="")
HistMeanMus(DivModel$mcmc, site=20, Pres = DataToJAGS$Present, main="Site 20", xlab="", ylab="")
HistMeanMus(DivModel$mcmc, site=32, Pres = DataToJAGS$Present, main="Site 32", xlab="", ylab="")
HistMeanMus(DivModel$mcmc, site=23, Pres = DataToJAGS$Present, main="Site 23", xlab="", ylab="")
mtext("Frequency", 2, outer=TRUE)
mtext("Species effect", 1, outer=TRUE)
```


## Beta Diversity

### Species Identity

#### Nestedness

We can calculate nestedness, and also look at the proportion of variance due to the gradient (@fig-nestedness).

```{r}
#| echo: false
#| label: fig-nestedness
#| fig-cap: "Nestedness. Variance on left, proportion due to the environment on the right. The horror that is the density on the right is real, and not due to the MCMC behaving poorly."
CalcNestedness <- function(mc, VarX=1, prop=FALSE) {
  resX <-  mc[,grep("mu.gam.z", colnames(mc))]^2*VarX
  res <- mc[,grep("sd.al.z", colnames(mc))]^2 + resX
  if(prop) {
    res.prop <- resX/res
    res <- cbind(res, res.prop)
  }
  as.mcmc(res)
}
Nestedness <- CalcNestedness(mc=DivPars, prop=TRUE)
# summary(Nestedness)
SummNest <- summary(Nestedness)

plot(Nestedness, trace = FALSE)

```

Most (`r round(100*SummNest$statistics["res.prop", "Mean"])`%) of the variation in nestedness is due to temperature. The effect has a posterior mean of `r round(Summ["mu.gam.z","Mean"], 2)` and a 95% credible interval of `r round(Summ["mu.gam.z","Lower95"], 2)` to `r round(Summ["mu.gam.z","Upper95"], 2)`. The distribution of the proporyion looks horrible here.


### Replacement

Replacement is $E_j \left(\text{Var}_i(\eta_{ij}^{(z)})\right) =  \sigma_{\alpha}^2 +  \left(\sigma^2_{\gamma^{(z)}} + (\bar{\gamma}^{(z)})^2\right) \text{Var}_i(x_i) + \pi^2/3$, and the density is plotted in @fig-replacement.

```{r}
#| echo: false
#| label: fig-replacement
#| fig-cap: "Densities of replacement: variance on left, proportion due to covariates on the right."
CalcReplacement <- function(mc, VarX=1, prop=FALSE) {
  resX <- (mc[,grep("sd.gam.z", colnames(mc))]^2 + mc[,grep("mu.gam.z", colnames(mc))]^2)*VarX
  res <- mc[,grep("sd.al.z", colnames(mc))]^2 + resX + pi/3
  if(prop) {
    res.prop <- resX/res
    res <- cbind(res, res.prop)
  }
  as.mcmc(res)
}
Replacement <- CalcReplacement(mc=DivPars, prop=TRUE)
SummReplacement <- summary(Replacement)
plot(Replacement, trace = FALSE)

```

Replacement is a number. Of it, about `r round(100*SummReplacement$statistics["res.prop", "Mean"])`% of the variation is due to temperature, so most of this variation cannot be explained by the gradient.

### Composition and GBH on GDM

Our compositional beta diversity statistic is $\sigma_{\alpha}^2 +  \left(\sigma^2_{\gamma^{(\lambda)}} + (\bar{\gamma}^{(\lambda)})^2\right) \text{Var}_i(x_i) + \sigma^2_\delta$

```{r}
#| echo: false
#| label: fig-CompBeta
#| fig-cap: "Densities of compositional diversity: variance on left, proportion due to the environment on right. The proportion due to the environment is equivalent to a statistic for GDM."
CalcBeta <- function(mc, VarX=1, prop=FALSE) {
  resX <- (mc[,grep("sd.gam.lam", colnames(mc))]^2+ mc[,grep("mu.gam.lam", colnames(mc))]^2)*VarX
  res <- mc[,grep("sd.al.lam", colnames(mc))]^2 + resX + mc[,grep("sdSiteOD", colnames(mc))]^2
  if(prop) {
    res.prop <- resX/res
    res <- cbind(res, res.prop)
  }
  as.mcmc(res)
}

Beta <- CalcBeta(mc=DivPars, prop=TRUE)
Summ.beta <-  summary(Beta)

plot(Beta, trace = FALSE)
```

We can see that we have beta diversity. For GDM fans, only about `r round(100*Summ.beta$statistics["res.prop", "Mean"])`% of the variation is driven by the temperature gradient.


## Gamma Diversity

Total Variation between sites & species. Does anyone care by this point? In case you do, it's in @fig-gamma.

```{r}
#| echo: false
#| label: fig-gamma
#| fig-cap: "Summary of posterior distribution of Gamma diversity (history of MCMC draws on left, density on right)."
Gamma <- as.mcmc.list(lapply(DivModel$mcmc, 
                             function(mc) {
                               res <- apply(mc[,grep("logmu", colnames(mc))], 1, var)
                               as.mcmc(res)
                             }))
plot(Gamma)

```

## For the Poster

```{r}
#| echo: false
#| label: fig-Poster
#| fig-height: 4
#| fig-cap: "Posterior distributions of diversity statistics. Posterior mode and 95% credible intervals"

CalcPosterStats <- function(mc, VarX=1, prop=FALSE) {
  
  sd.al.lam <- mc[,grep("sd.gam.lam", colnames(mc))]
  gamma.lam <- mc[,grep("gamma.lam", colnames(mc))]
  sd.be.lam <- mc[,grep("sd.be.lam", colnames(mc))]
  sd.al.z <- mc[,grep("sd.al.z", colnames(mc))]
  gamma.z <- mc[,grep("gamma.z", colnames(mc))]
  sdSiteOD <- mc[,grep("sdSiteOD", colnames(mc))]
  sd.be.z <- mc[,grep("sd.be.z", colnames(mc))]
  
  Stats.l <- cbind(
  Replacement.l = sd.al.lam^2 + apply(gamma.lam, 1, mean)^2 + apply(gamma.lam, 1, sd)^2 +
    sdSiteOD^2,
  Nestedness.l = sd.al.lam^2 + apply(gamma.lam, 1, mean)^2,
  Disparity.l = sd.be.lam,
  Unevenness.l = sd.be.lam^2 + sdSiteOD^2
  )

  Stats.z <- cbind(
  Replacement.z = sd.al.z^2 + apply(gamma.z, 1, mean)^2 + apply(gamma.z, 1, sd)^2 + (pi^2)/3,
  Nestedness.z = sd.al.z^2 + apply(gamma.z, 1, mean)^2, 
  Disparity.z = sd.be.z,
  Unevenness.z = sd.be.z^2 + (pi^2)/3
  )

  if(prop) {
    Replacement.l.x = (apply(gamma.lam, 1, mean)^2 + apply(gamma.lam, 1, sd)^2)
    Nestedness.l.x = apply(gamma.lam, 1, mean)^2
    Stats.l.prop <- cbind(
      Replacement.l.p = Replacement.l.x/Stats.l[,"Replacement.l"], 
      Nestedness.l.p = Nestedness.l.x/Stats.l[,"Nestedness.l"]
    )
    
    Replacement.z.x = (apply(gamma.z, 1, mean)^2 + apply(gamma.z, 1, sd)^2)
    Nestedness.z.x = apply(gamma.z, 1, mean)^2
    Stats.z.prop <- cbind(
      Replacement.z.p = Replacement.z.x/Stats.z[,"Replacement.z"], 
      Nestedness.z.p = Nestedness.z.x/Stats.z[,"Nestedness.z"]
    )

    
    # Stats.z.prop <- c(
    #   Replacement.z.p = (apply(gamma.z, 1, mean)^2 + apply(gamma.z, 1, sd))^2/Stats.z[,"Replacement.z"], 
    #   Nestedness.z.p = apply(gamma.lam, 1, mean)^2/Stats.z[,"Nestedness.z"]
    # )
    Stats.l <- cbind(Stats.l, Stats.l.prop)
    Stats.z <- cbind(Stats.z, Stats.z.prop)
  }
  
  res <- cbind(Stats.l, Stats.z)
  as.mcmc(res)
}


thing <- as.mcmc.list(lapply(DivModel$mcmc, CalcPosterStats, prop=TRUE))
summ.stats <- summary(thing)
Nop <- !grepl("p$", rownames(summ.stats$statistics))
At <- rep(1:4,2) + rep(c(-0.1,0.1), each=4)

par(mar=c(4.1,8,1,1))
plot(summ.stats$statistics[Nop,"Mean"],At, cex=1, pch=15, xlim=range(summ.stats$quantiles),
     yaxt="n", col=rep(c(2,3), each=4), xlab="Diversity Statistic", ylab="")
segments(summ.stats$quantiles[Nop,"2.5%"],At, summ.stats$quantiles[Nop,"97.5%"],At, 
         col=rep(c(2,3), each=4))
axis(2, c("Replacement", "Nestedness", "Disparity", "Unevenness"), at=1:4, las=1)
legend(6.3,3,c("Abundance", "Incidence"), fill=2:3, cex=1)
```

```{r}
#| echo: false
#| label: fig-PosterProps
#| fig-height: 4
#| fig-cap: "Posterior distributions for proportions of diversity statistics explained by temperature. Posterior mode and 95% credible intervals"


Nop <- !grepl("p$", rownames(summ.stats$statistics))
At <- rep(1:2,2) + rep(c(-0.1,0.1), each=2)

par(mar=c(4.1,8,1,1))
plot(summ.stats$statistics[!Nop,"Mean"],At, cex=1, pch=15, xlim=c(0,0.2),
     yaxt="n", col=rep(c(2,3), each=2), xlab="Proportion of Diversity", ylab="")
segments(summ.stats$quantiles[!Nop,"2.5%"],At, summ.stats$quantiles[!Nop,"97.5%"],At, 
         col=rep(c(2,3), each=2))
axis(2, c("Replacement", "Nestedness"), at=1:2, las=1)
legend(0.1,1.7,c("Abundance", "Incidence"), fill=2:3, cex=1)
```



<!---
## Ignore this bit.

Finally, some code to find out which covariate to use. This is not run when the document is rendered. It suggest that there are positive effects of silt/clay chlorophyll a and elevation on abundance, and possibly a negative effect of temperature on presence.

```{r}
#| eval: false
#| echo: false
EnvToUse <- c("TOC", "elevation", "temperature", "DIN", "RDP", 
                         "Chl.a", "silt_clay")
EnvSc <- scale(Env[,EnvToUse])
DataToJAGS$Env <- as.matrix(EnvSc)
DataToJAGS$NEnv <- ncol(EnvSc)
# DataToJAGS$Island <- Counts$Island
# DataToJAGS$Transect <- Counts$Transect
# DataToJAGS$Station <- Counts$Station
# DataToJAGS$Season <- Counts$Season
# pairs(Env[,9:16])
# 
# par(mfrow=c(2,4))
# apply(Env[,9:16], 2, hist)

cat("
model {
  for(sp in 1:NSp) {
    for(rep in 1:NRep) {
      Abund[rep, sp] ~ dpois(lambda[rep, sp])
      lambda[rep, sp] <- mu[rep, sp]*Present[Site[rep], sp]
      OD[rep, sp] ~ dnorm(0, tauOD)
      log(mu[rep, sp]) <- logmu[Site[rep], sp] + OD[rep, sp]
    }
    for(site in 1:NSite) {
      Present[site, sp] ~ dbern(p[site, sp])
      logmu[site, sp] <- mu.lam + al.lam[site] + be.lam[sp] + 
          inprod(Env[site,], mu.gam.lam[])
      logit(p[site, sp]) <- mu.z + al.z[site] + be.z[sp] + 
          inprod(Env[site,], mu.gam.z[])
    }
    be.lam[sp] ~ dnorm(mu.be.lam, tau.be.lam)
    be.z[sp] ~ dnorm(mu.be.z, tau.be.z)

#    for(env in 1:NEnv) {
#      gamma.z[sp,env] ~ dnorm(mu.gam.z[env], tau.gam.z[sp,env])
#      gamma.lam[sp,env] ~ dnorm(mu.gam.lam[env], tau.gam.lam[sp,env])
#      logtau.gam.z[sp,env] ~ dunif(-10,10)
#      tau.gam.z[sp,env] <- exp(logtau.gam.z[sp,env])
#      logtau.gam.lam[sp,env] ~ dunif(-10,10)
#      tau.gam.lam[sp,env] <- exp(logtau.gam.lam[sp,env])
#    }
  }
  for(site in 1:NSite) {
    al.lam[site] ~ dnorm(mu.al.lam, tau.al.lam)
    al.z[site] ~ dnorm(mu.al.z, tau.al.z)
    SR[site] <- sum(Present[site, ])
  }
  for(env in 1:NEnv) {
      mu.gam.z[env] ~ dnorm(0,1)
      mu.gam.lam[env] ~ dnorm(0,1)
  }

  mu.be.lam ~ dnorm(0, 1)
  mu.be.z ~ dnorm(0, 1)
  mu.al.lam ~ dnorm(0, 1)
  mu.al.z ~ dnorm(0, 1)
  mu.lam ~ dnorm(0, 1)
  mu.z ~ dnorm(0, 1)

  Mean.lam <- mu.lam + mean(al.lam[]) + mean(be.lam[])
  Mean.z <- mu.z + mean(al.z[]) + mean(be.z[])
  
  tau.be.lam <- pow(sd.be.lam, -2); sd.be.lam ~ dexp(1)
  tau.be.z <- pow(sd.be.z, -2); sd.be.z ~ dexp(1)
  tau.al.lam <- pow(sd.al.lam, -2); sd.al.lam ~ dexp(1)
  tau.al.z <- pow(sd.al.z, -2); sd.al.z ~ dexp(1)

  tauOD <- pow(sdOD, -2); sdOD ~ dexp(1)


}", file="BeatingBetaVarSelecCode.jag")

GetMoreInits <- function(data) {
  init <- GetInits(data)
#  init$logtau.gam.z <- matrix(-2, nrow=data$NSp, ncol=data$NEnv)
#  init$logtau.gam.lam <- matrix(-2, nrow=data$NSp, ncol=data$NEnv)

  init$mu.gam.z <- rnorm(data$NEnv, 0, 0.2)
  init$mu.gam.lam <- rnorm(data$NEnv, 0, 0.2)
  init
}
try <- GetMoreInits(DataToJAGS)
Inits <- replicate(2, GetMoreInits(DataToJAGS), simplify = FALSE)

ToMonitor.gams <- c("mu.gam.z", "mu.gam.lam")

DivModelSelec <- run.jags(model = "BeatingBetaVarSelecCode.jag", 
                          monitor=ToMonitor.gam, data=DataToJAGS,
                  inits = Inits, method="parallel", # silent.jags=TRUE,
         n.chains = 2, adapt = 5e3, burnin = 5e3, sample = 1e3, thin = 10)
plot(DivModelSelec, plot.type = "trace")
SummSel <- summary(DivModelSelec)
LamSel <- grep("lam", rownames(SummSel))
ZSel <- grep("z", rownames(SummSel))

par(mfrow=c(1,2), mar=c(2,1,4,1), oma=c(2,6,2,2))
plot(SummSel[ZSel,"Mean"], ZSel, 
     xlim=c(min(SummSel[ZSel,"Lower95"]), max(SummSel[ZSel,"Upper95"])),
     xlab="Coefficient", ylab="", yaxt="n", main="Presence")
segments(SummSel[ZSel,"Lower95"], ZSel, SummSel[ZSel,"Upper95"], ZSel)
axis(2, at=ZSel, labels=EnvToUse, las=1)
abline(v=0, lty=2)

plot(SummSel[LamSel,"Mean"], LamSel, 
     xlim=c(min(SummSel[LamSel,"Lower95"]), max(SummSel[LamSel,"Upper95"])),
     xlab="Coefficient", ylab="", yaxt="n", main="Abundance")
segments(SummSel[LamSel,"Lower95"], LamSel, SummSel[LamSel,"Upper95"], LamSel)
# axis(2, at=LamSel, labels=EnvToUse, las=1)
abline(v=0, lty=2)


```
--->@


Swapping sites and soecies

"alpha diversity"

Variation between sites for one species., Mean: rarity/commonnness. Variance: specialisation

Need z & lambda together to understand patterns. Should write them out.


"beta diversity": environmental variance. Resilinace?

Need to add traits explicitly. alpha_i Tr_j effect: what does it mean?

