---
title: "Better GDM"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
editor: source
bibliography: references.bib
execute: 
  cache: true
---

## The Problem

The idea of GDM, modelling matrices of dissimilarity, was suggested by @Ferrier2002. Recently, @WhiteMEE2023 suggested improvements to generalized dissimilarity modelling. Here we will develop an alternative, based on modelling the data rather than the distances between data.

The ecological problem is to explain how beta diversity, i.e. variation between sites, is affected by predictors, such as the environment. This can be done using GDM, which regresses distances between communities against the environemntal covariates. Here we suggest a more direct approach: model the effects of the environment on the species, and then look at how this affects variation in the communities.

## The Model

We assume we have a site by species matrix, $Y$, with each row being observations of species (abundance, presentece etc.), and each column being a site. So $Y_{ij}$ is the observation for the $i^th$ species ($i=1,\dots, S$) and $j^th$ species ($j=1,\dots, N$). For each site we have $P$ covariates, $x_{jk}$ ($k=1,\dots,P$), so these are in a $N \times P$ matrix $X$.

We assume we model this data using a GLM, i.e. we assume $Y_{ij}$ follows some distribution in the exponential family, with expected value $\mu = g(\eta)$, where $g()$ is a link function, and $\eta$ is a linear predictor:

$$
\eta_{ij} = \alpha_i + \kappa_j + \beta_{ij}
$$

For identifiability we assume $\sum_j \kappa_j = 0$ and $\sum_j \beta_j = 0$

So $\alpha_i$ is the mean commonness of a species, and $\kappa$ is the abundance at a site: we will assume this is not interesting, as it is affected by the sampling process. $\beta_{ij}$ is where beta diversity lies (at least compositional beta diversity). The total beta diversity can be measured as $\text{Var}(\beta) = \sigma^2_\beta$. We can clearly model $\beta_{ij}$ further, e.g. for a single covariate $x_j$

$$
\beta_{ij} = \beta_0 + \delta_i x_j + \epsilon_{ij}
$$

i.e. each species has its own response to $x$.

Then,

$$
\begin{aligned}
\text{Var}(\beta) &= \text{Var}(\beta_0 + \delta_i x_j + \epsilon_{ij}) \\
&= \frac{1}{(S-1)(N-1)}\sum_{i=1}^S \sum_{j=1}^N (\delta_i (x_j - \bar{x}) + \epsilon_{ij})^2 \\
&= \frac{1}{(S-1)(N-1)} \left(\sum_{i=1}^S \sum_{j=1}^N \delta_i^2 (x_j - \bar{x})^2 + \sum_{i=1}^S \sum_{j=1}^N\epsilon_{ij}^2 \right)\\
&= \frac{1}{(S-1)} \sum_{i=1}^S \delta_i^2 \frac{1}{(N-1)}\sum_{j=1}^N  (x_j - \bar{x})^2 + \frac{1}{(S-1)(N-1)}\sum_{i=1}^S \sum_{j=1}^N\epsilon_{ij}^2 \\
&= \frac{1}{(S-1)} \sum_{i=1}^S \delta_i^2 \text{Var}(x) + \text{Var}(\epsilon_{ij})\\
&= \frac{\text{Var}(x)}{(S-1)} \sum_{i=1}^S (\delta_i - \bar{\delta} + \bar{\delta})^2 + \text{Var}(\epsilon_{ij})\\
&= \frac{\text{Var}(x)}{(S-1)} \sum_{i=1}^S (\delta_i - \bar{\delta})^2 - S \bar{\delta}^2 + \text{Var}(\epsilon_{ij})\\
&= \text{Var}(\delta_i) \text{Var}(x_j) + \frac{S}{S-1} \bar{\delta}^2 \text{Var}(x) + \text{Var}(\epsilon_{ij})
\end{aligned}
$$

because $\sum_{i}(\delta_i - \bar{\delta} + \bar{\delta})^2 = \sum_{i}(\delta_i - \bar{\delta})^2 + \sum_{i}(\delta_i - \bar{\delta})\bar{\delta} + \sum_{i}\bar{\delta})^2 = \sum_{i}(\delta_i - \bar{\delta})^2 + \bar{\delta} (\sum_{i}\delta_i - S\bar{\delta}) + S\bar{\delta}^2 = \sum_{i}(\delta_i - \bar{\delta})^2 + S\bar{\delta}^2$.

So, we can model the species abundance as a function of the environment. We can extend this to more than one covariate, if they are assumed to have independent effects between species (i.e. $\delta_{ik}$ is independent of $\delta_{il}$ for species $i$ and covariates $k$ and $l$).

## Relationship to Distance

If we have two vectors of biodiversity, $\boldsymbol{\eta}_1$ and $\boldsymbol{\eta}_2$ then the distance between then the distance between them is just the Euclidean distance:

$$
D_{12} = \lVert\boldsymbol{\eta}_1 - \boldsymbol{\eta}_2 \rVert =\sqrt{\sum_{i=1}^S (\eta_{i1} - \eta_{i2})^2}
$$

So that

$$
\begin{aligned}
D^2_{12} &= \sum_{i=1}^S (\eta_{i1} - \eta_{i2})^2 \\
&= \sum_{i=1}^S \left(\alpha_i + \kappa_1 + \beta_{i1} - (\alpha_i + \kappa_2 + \beta_{i2})\right)^2 \\
&= \sum_{i=1}^S \left((\kappa_1 - \kappa_2) + (\beta_{i1} - \beta_{i2})\right)^2 \\
&= \sum_{i=1}^S \left((\kappa_1 - \kappa_2)^2 + (\kappa_1 - \kappa_2)(\beta_{i1} - \beta_{i2}) + (\beta_{i1} - \beta_{i2})^2\right) \\
&= S (\kappa_1 - \kappa_2)^2 + (\kappa_1 - \kappa_2) \sum_{i=1}^S (\beta_{i1} - \beta_{i2}) + \sum_{i=1}^S(\beta_{i1} - \beta_{i2})^2\\
&= S (\kappa_1 - \kappa_2)^2 +  \sum_{i=1}^S(\beta_{i1} - \beta_{i2})^2\\
&= S (\kappa_1 - \kappa_2)^2 +  \sum_{i=1}^S\beta_{i1}^2 - 2\sum_{i=1}^S\beta_{i1}\beta_{i2} + \sum_{i=1}^S\beta^2_{i1}\\
&= S (\kappa_1 - \kappa_2)^2 +  (S-1) \left(\text{Var}(\beta_{\cdot 1}) + \text{Var}(\beta_{\cdot 2}) - 
2\text{Cov}(\beta_{\cdot 1} \beta_{\cdot 2})\right)
\end{aligned}
$$

Where the $(\kappa_1 - \kappa_2) \sum_{i=1}^S (\beta_{i1} - \beta_{i2})$ term disappears because, by construction, $\sum_{i=1}^S \beta_{ij}=0$. $\text{Var}_{ij}(X)$ is the variance of $X$ in populations $i$ and $j$.

If we are interested in compositional diversity, we can ignore the terms involving the $\kappa$s, so the distance is just a function of the second moments of $\beta$.

## Examples

We use the three examples from @WhiteMEE2023. First we need to do some preliminary work: importing packages, defining some terms (e.g. priors), and several functions that will b eused later.

```{r Bookkeeping}
#| results: hide
#| output: false

library(zen4R) # used to download SA data
library(INLA)
library(gdm)
library(taxize)

# Define PC priors, to stop the models exploding horribly
PCHyper <- list(prec = list(prior="pc.prec", param=c(u=0.5, alpha=0.001)))
PCHyper2 <- list(prec = list(prior="pc.prec", param=c(u=1, alpha=0.1)))

# Number of simulations from posterior to calculate statistics
nsim <- 1000

# Function to calculate variance components
CalcPostVar <- function(INLAsamp, data) {
  # Extract main effects of covariates
  LatentNames <- attr(INLAsamp$latent, "dimnames")[[1]]
  RFnames <- paste(c(gsub("Precision for ", "", names(INLAsamp$hyperpar)), 
                     "(Intercept)", "Predictor"), collapse="|")
  FxNames <- LatentNames[!grepl(RFnames, LatentNames)]
  RFNames <- gsub(":.", "", FxNames)
  # Calculate/extract parts of variance terms
  VarHyper <- 1/INLAsamp$hyperpar
  wh.delta <- sapply(RFNames, function(wh, nm) grep(wh, nm), nm=names(VarHyper))
  VarDelta <- VarHyper[wh.delta]
  VarOth <- VarHyper[!seq_along(VarHyper)%in%wh.delta]
  
  VarX <- var(data[, RFNames])
  S <- nlevels(data$Species)

  # Deltabar
  Fx.wh <- which(attr(INLAsamp$latent, "dimnames")[[1]]%in%FxNames)
  Deltabar <- INLAsamp$latent[Fx.wh]
  names(Deltabar) <- attr(INLAsamp$latent, "dimnames")[[1]][Fx.wh]
  
  
  AllVars <- c(MeanDelta = t(Deltabar)%*%VarX%*%Deltabar,
               VarDelta = sum(VarDelta*diag(VarX)), # not sure this is right
               VarOth)
  return(AllVars)
}

# Function to transform log precision of hyper parameters to standard deviation
CalcHyperSDs <- function(mod, use=NULL) {
  if(is.null(use)) use <- seq_along(mod$internal.marginals.hyperpar)
  
  sd.lst <- lapply(mod$internal.marginals.hyperpar[use], function(lst) {
    marg <- inla.tmarginal(function(x) exp(-x/2), lst)
    res <- c(inla.mmarginal(marg), inla.hpdmarginal(c(0.5, 0.95), marg))
    names(res) <- c("mode", "Lower50", "Lower95", "Upper50", "Upper95")
    res
  })
  sd.df <- as.data.frame(do.call(rbind, sd.lst))
  sd.df
}

# Function to create linear combinations of fixed & random effects for variable var
CreateLinComb <- function(var, Levels, prefix="Species.") {
  lc <- inla.make.lincombs(var= rep(1, nlevels(Levels)),
                           tt = diag(1, nlevels(Levels)))
  lc <- lapply(lc, function(l, var) {
    names(l[[1]]) <- var
    names(l[[2]]) <- paste0(prefix, var)
    l
  }, var=var)
  names(lc) <- paste0(var, names(lc))
  lc
}


# Function to plot fixed effects
# PlotFixed(fixed=SAFamSumm$fixed, use=Use.fixed)

PlotFixed <- function(fixed, use) {
  plot(fixed[use, "mean"], use, yaxt="n", xlab="Mean Effect", ylab="", 
       xlim=range(fixed[use, c("0.025quant", "0.975quant")]),
       main="Mean Effect")
  segments(fixed[use, "0.025quant"], use, fixed[use, "0.975quant"], use)
  axis(2, gsub("\\.sc", "", rownames(fixed)[use]), at=use, las=1)
  abline(v=0, lty=2)
}

# Function to plot hyperparameter SDs
# PlotHyper(SDs=SAHyperSDs, use=Use.hyper)
PlotHyper=function(SDs, use=NULL) {
  if(is.null(use)) use <- 1:nrow(SDs)
  plot(SDs[use,"mode"], use, yaxt="n", xlab="Std Dev", ylab="", 
       xlim=range(c(0,SDs[,c("Lower95", "Upper95")])),
       main="Std Dev across Species")
  segments(SDs[use,"Lower50"], use, SDs[use,"Upper50"], use, lwd=3, lend=2)
  segments(SDs[use,"Lower95"], use, SDs[use,"Upper95"], use)
}

# Function to plot random effects and their means
PlotEff <- function(summ, mean, ...) {
  At <- rank(summ$mean)
  plot(summ$mean, At, yaxt="n", type="n", ...)
  rect(mean["0.025quant"], -10, mean["0.975quant"], 1e4, col="pink", border=NA)
  abline(v=mean["mean"], col=2)
  points(summ$mean, At)
  segments(summ[,"0.025quant"], At, 
           summ[,"0.975quant"], At)
  abline(v=0, lty=2)
}

```

### Example 1: South African flora

This is data with 288 species and 413 sites. The data is given in wide format, so we need to convert it to long. The data can be downloaded from [Zenodo](https://zenodo.org/records/10091442).

```{r GetDataSA}
#|output: false

# Download the data, if it hasn't already been done. Cleans up afterwards, too 
if(!file.exists("data/sa_species_data.csv")) {
  download_zenodo("10.5281/zenodo.10091441", path="data")
  unzip("data/spGDMM-code-spGDMM_v1.zip", exdir="data")
  DataFiles <- dir("data/philawhite-spGDMM-code-d8279ee/data/")
  sapply(DataFiles, function(x) {
    cat(x, file.exists(file.path("data/philawhite-spGDMM-code-d8279ee/data/",x)), "\n")
    file.copy(from=file.path("data/philawhite-spGDMM-code-d8279ee/data/",x),
              to=file.path("data/",x), overwrite = TRUE)
  })
    unlink("data/philawhite-spGDMM-code-d8279ee/", recursive = TRUE)
    unlink("data/spGDMM-code-spGDMM_v1.zip", recursive = TRUE)
}

Data <- read.csv("data/sa_species_data.csv")

# SpNames <- gsub("\\.", " ", names(Data)[grep("\\.", names(Data))])
SpNames <- names(Data)[grep("\\.", names(Data))]

SpData.wide <- Data[, SpNames]
IsZero <- apply(SpData.wide, 2, function(x) mean(x==0))

UseNames <- names(IsZero)[IsZero<0.95]
RemoveNames <- SpNames[!SpNames%in%UseNames]
NamesToLong <- names(Data)[!names(Data)%in%RemoveNames]

SAData.long <- reshape(Data[,NamesToLong], direction = "long", varying = UseNames,
                       v.names = "Percent", timevar="SpeciesID")
SAData.long$Species <- SpNames[SAData.long$SpeciesID]
SAData.long$Percent2 <- pmax(0.1, SAData.long$Percent)
SAData.long$Abund <- log(SAData.long$Percent2/(100-SAData.long$Percent2))

SA.EnvCov <- c("gmap", "RFL_CONC", "HeatLoadIndex30m", "tmean13c", 
            "SoilConductivitymSm", "SoilTotalNPercent")

ScCov <- apply(SAData.long[,SA.EnvCov], 2, scale)
colnames(ScCov) <- paste0(colnames(ScCov), ".sc")

SAData <- cbind(SAData.long, ScCov)

# SAData$Species <- factor(SAData$Species)
SAData$plot <- factor(SAData$plot)

```

Covariates: 

- gmap: mean annual precipitation 
- RFL_CONC: rainfall concentration 
- Elevation30m: elevation (at 30m resolution) 
- HeatLoadIndex30m: heat load index (at 30m resolution) 
- tmean13c: mean annual temperature 
- SoilConductivitymSm: soil conductivity 
- SoilTotalNPercent: Nitrogen concentration in soil (%)

We remove elevation because of the high correlation with tmean13c (`r round(cor(SAData.long$tmean13c, SAData.long$Elevation30m), 2)`), and all species that occur in \<1% of sites, giving `r length(UseNames)` species out of the original 288.

We fit the model with INLA, and use relatively tight PC priors to control the estimates of the environmental effects (without doing this everything is horribly unstable, and even doing this the model is not very stable: slight changes change the effects of covariates).

```{r, FitModelSA}
#| warning: false
# Create formula for INLA
EnvForm <- sapply(SA.EnvCov, function(var) {
  paste0(var, ".sc + f(Species.", var, ".sc, ", var, 
         ".sc, hyper = PCHyper)")
})

SA.Formula <- formula(
  paste0("Abund ~ f(Species, hyper = PCHyper) + f(plot, hyper = PCHyper) + ", 
         paste(EnvForm, collapse=" + "))
)

SAenvInd <- data.frame(sapply(SA.EnvCov, function(nm, dat)
  assign(nm, dat), dat=SAData$Species))
names(SAenvInd) <- paste0("Species.", names(SAenvInd), ".sc")

SAData <- cbind(SAData, SAenvInd)

LCs.lst <- lapply(paste0(SA.EnvCov, ".sc"), CreateLinComb,
                  Levels=factor(SAData$Species))
SA.LCs <- do.call(c, LCs.lst)

# int.strategy = "ccd" or int.strategy = "grid"
# Fit the model
SAinla <- inla(SA.Formula, data=SAData, control.compute=list(config = TRUE), 
               control.inla=list(int.strategy = "ccd"), lincomb = SA.LCs)
SASumm <-summary(SAinla)

```

From this we first need to calculate the statistics described above. Because the statistics are not linear, we need to simulate from the posterior. There is an approximation here because INLA will only simulate from the integration points of the hyperparameters, not from their full distribution. This might be fixable later.

```{r, CalcStatsSA}
#| eval: true
#| fig-cap: Posterior distributions of diversity components
#| fig-height: 3

SAsamps <- inla.posterior.sample(nsim, SAinla)

SAPosts <- data.frame(t(sapply(SAsamps, CalcPostVar, data=SAData)))
Use <- !names(SAPosts)%in%paste0("Precision.for.", c("Species", "plot"))
SA.PropVar <- sweep(SAPosts[,Use], 1, rowSums(SAPosts[,Use]), "/")
names(SA.PropVar) <- gsub("Precision.for.the.Gaussian.observations", "VarEps", 
                          names(SA.PropVar), fixed=TRUE)

par(mfrow=c(1,3))
plot(density(SA.PropVar$MeanDelta), main="Mean Covariate")
plot(density(SA.PropVar$VarDelta), main="Variation in Covariate")
plot(density(SA.PropVar$VarEps), main="Residual")

# Summary
SA.SummVars <- cbind(
  Mode=MCMCglmm::posterior.mode(coda::as.mcmc(SA.PropVar)),
  coda::HPDinterval(coda::as.mcmc(SA.PropVar), prob = 0.95)
)

```

We find that `r round(100*(1-SA.SummVars["VarEps", "Mode"]), 2)`% of the variation in beta diversity is explained by the covariates, with a 95% HPDI of `r round(100*(1-SA.SummVars["VarEps", "upper"]), 2)`% - `r round(100*(1-SA.SummVars["VarEps", "lower"]), 2)`%. So only a small fraction of the variation is due to the environment. The distribution for the variation in the covariate looks horrible because it is only drawn from the integration points, as noted above. This makes it a bit of a mess, but doesn't change the overall message in any of the examples.

We can look at the importance of the variables. First, by plotting the mean effects and their standard deviations (across species). Then we can look at their contributions to the variance.

```{r PlotMeansSA}
#| eval: true
#| warning: false
# Plot means
Use.fixed <- which(!grepl("\\(Int", rownames(SASumm$fixed)))
Use.hyper <- sapply(SA.EnvCov, function(x, y) grep(x, y),
                    y=names(SAinla$marginals.hyperpar))

SAHyperSDs <- CalcHyperSDs(SAinla, use=Use.hyper)

par(mar=c(4.1,1,3,1), oma=c(1,8,1,1), mfrow=c(1,2))
 PlotFixed(fixed=SASumm$fixed, use=Use.fixed)
 PlotHyper(SDs=SAHyperSDs, use=NULL)

```

We can see that temperature has the largest effect, both in the average effect and the variation across species. Four of the effects appear not to have any variation between species. But, as noted above, this does not seem to be stable: change hte model or priors and we get slightly different results.

Finally, we can visualise the effects of the environment on each species. We can clearly see the effect of temperature, and that different species prefer it warmer or colder. But we can see this across other covariates too.

```{r PlotCovarEffsSA}
#| fig-height: 8
#| results: hide

SA.RanEff <- SAinla$summary.lincomb.derived
SA.UseSumRan <- grep("Species\\.", names(SAinla$summary.random))
Xlims <- range(SA.RanEff[,c("0.025quant", "0.975quant")])

par(mar=c(2.1,1,3,1), mfrow=c(2,3))
sapply(SA.EnvCov, function(env, RanEff, SummFix, ...) {
PlotEff(summ=RanEff[grep(env, rownames(RanEff)),], 
        mean=SummFix[paste0(env, ".sc"),],
        ylab="", xlab="", main=env, ...)
}, RanEff=SA.RanEff, SummFix=SASumm$fixed, xlim=Xlims)

```

### Example 1a: South African flora at the family level

Like @WhiteMEE2023, we will also look at this dataat the family level. This means we have 52 families and 413 sites. We will compare the family and species level analyses with each other and with @WhiteMEE2023 below. First, download and format the data.

```{r GetDataSAFam}
#|output: false

FamData <- read.csv("data/sa_family_data.csv")
FamNames <- names(FamData)[!names(FamData)%in%names(Data)]
FamData.wide <- FamData[, FamNames]
IsZero <- apply(FamData.wide, 2, function(x) mean(x==0))

UseNames <- names(IsZero)[IsZero<0.99]
RemoveNames <- FamNames[!FamNames%in%UseNames]
NamesToLong <- names(FamData)[!names(FamData)%in%RemoveNames]

FamData.long <- reshape(FamData[,NamesToLong], direction = "long", 
                        varying = UseNames,
                       v.names = "Percent", timevar="FamilyID")

FamData.long$Species <- FamNames[FamData.long$FamilyID]
FamData.long$Percent2 <- pmax(0.1, FamData.long$Percent)
FamData.long$Abund <- log(FamData.long$Percent2/(100-FamData.long$Percent2))

Fam.EnvCov <- c("gmap", "RFL_CONC", "HeatLoadIndex30m", "tmean13c", 
            # "Elevation30m", 
            "SoilConductivitymSm", "SoilTotalNPercent")

ScCov <- apply(FamData.long[,Fam.EnvCov], 2, scale)
colnames(ScCov) <- paste0(colnames(ScCov), ".sc")

SAFamData <- cbind(FamData.long, ScCov)

SAFamData$Family <- factor(UseNames[SAFamData$FamilyID])
SAFamData$plot <- factor(SAFamData$plot)

```

The covariates are as before, as is the use of INLA and the priors:

```{r, FitModelSAFam}
#| warning: false
# Create formula for INLA
SAFam.Formula <- formula(
  paste0("Abund ~ f(Family, hyper = PCHyper) + f(plot, hyper = PCHyper) + ", 
         paste(
           c(paste0(Fam.EnvCov, ".sc"), 
             sapply(paste0(Fam.EnvCov, ".sc"), function(eu) {
               paste0("f(Family.", eu, ", ", eu, ", hyper = PCHyper)")
             })), collapse=" + "))
)

SAFamenvInd <- data.frame(sapply(Fam.EnvCov, function(nm, dat) assign(nm, dat), 
                              dat=SAFamData$Family))
names(SAFamenvInd) <- paste0("Family.", names(SAFamenvInd), ".sc")
SAFamData <- cbind(SAFamData, SAFamenvInd)

# Linear combinations
LCs.lst <- lapply(paste0(SA.EnvCov, ".sc"), CreateLinComb,
                  Levels=factor(SAFamData$Family), prefix="Family.")
SAFam.LCs <- do.call(c, LCs.lst)

# Fit the model
SAFaminla <- inla(SAFam.Formula, data=SAFamData, 
                  control.compute=list(config = TRUE), 
               control.inla=list(int.strategy = "ccd"), lincomb = SAFam.LCs)
SAFamSumm <-summary(SAFaminla)

```

As before, we now calculate the statistics.

```{r, CalcStatsSAFam}
#| eval: true
#| fig-height: 3

SAFamsamps <- inla.posterior.sample(nsim, SAFaminla)

SAFamPosts <- data.frame(t(sapply(SAFamsamps, CalcPostVar, data=SAFamData)))
Use <- !names(SAFamPosts)%in%paste0("Precision.for.", c("Family", "plot"))
SAFam.PropVar <- sweep(SAFamPosts[,Use], 1, rowSums(SAFamPosts[,Use]), "/")
names(SAFam.PropVar) <- gsub("Precision.for.the.Gaussian.observations", "VarEps", 
                          names(SAFam.PropVar), fixed=TRUE)

par(mfrow=c(1,3))
plot(density(SAFam.PropVar$MeanDelta), main="Mean Covariate")
plot(density(SAFam.PropVar$VarDelta), main="Variation in Covariate")
plot(density(SAFam.PropVar$VarEps), main="Residual")

# Summary
SAFam.SummVars <- cbind(
  Mode=MCMCglmm::posterior.mode(coda::as.mcmc(SAFam.PropVar)),
  coda::HPDinterval(coda::as.mcmc(SAFam.PropVar), prob = 0.95)
)

```

We find that `r round(100*(1-SAFam.SummVars["VarEps", "Mode"]), 2)`% of the variation in beta diversity is explained by the covariates, with a 95% HPDI of `r round(100*(1-SAFam.SummVars["VarEps", "upper"]), 2)`% - `r round(100*(1-SAFam.SummVars["VarEps", "lower"]), 2)`%, so a bit more of the beta diversity at the family level is explained by the environmental covariates than at the species level. We can look at the individual covariates:

```{r PlotEffectsFam}

Use.fixed <- which(!grepl("\\(Int", rownames(SAFamSumm$fixed)))
Use.hyper <- sapply(SA.EnvCov, function(x, y) grep(x, y),
                    y=names(SAFaminla$marginals.hyperpar))

SAFamHyperSDs <- CalcHyperSDs(SAFaminla, use=Use.hyper)

par(mar=c(4.1,1,3,1), oma=c(1,8,1,1), mfrow=c(1,2))
 PlotFixed(fixed=SAFamSumm$fixed, use=Use.fixed)
 PlotHyper(SDs=SAFamHyperSDs, use=NULL)

```

Again, temperature has the largest effect, both in the average effect and the variation across species. But now the estiamtes look happier - nothing going to zero. Finally, some more detailed plots:

```{r PlotCovarEffsFam}
#| fig-height: 8
#| results: hide

SA.RanEff <- SAinla$summary.lincomb.derived
SA.UseSumRan <- grep("Species\\.", names(SAinla$summary.random))
Xlims <- range(SA.RanEff[,c("0.025quant", "0.975quant")])

SAFam.RanEff <- SAFaminla$summary.lincomb.derived
UseSumRan <- grep("Family\\.", names(SAFaminla$summary.random))
Xlims <- range(unlist(lapply(SAFaminla$summary.random[UseSumRan], function(lst) 
  range(lst[,c("0.025quant", "0.975quant")]))))


par(mar=c(2.1,1,3,1), mfrow=c(2,3))
sapply(SA.EnvCov, function(env, RanEff, SummFix, ...) {
PlotEff(summ=RanEff[grep(env, rownames(RanEff)),], 
        mean=SummFix[paste0(env, ".sc"),],
        ylab="", xlab="", main=env, ...)
}, RanEff=SAFam.RanEff, SummFix=SAFamSumm$fixed, xlim=Xlims)

```


#### Comparison with spGDMM

With spGDMM, @WhiteMEE2023 could compare models using cross-validation, were able to compare the effects of the different covariates, and were qualitatively able to compare the covariate effects with a spatial term: 

      The scale of the differences in $\eta(\mathbf{s}, \mathbf{s}')$ is large 
      relative to the scale of the scaled warping functions in Figure 4. This 
      suggests that the spatial random effect contributes to the mean dissimilarity 
      more than the environmental covariates, even when considering the additive 
      effect of many environmental variables. 

Here we don't (yet) have a spatial term, but we can compare the covariate effects to the total variance in diversity, and see that at the species level almost all of the variation (`r round(100*(SAFam.SummVars["VarEps", "Mode"]), 2)`%) is unexplained, whilst at the family level about `r round(100*(1-SAFam.SummVars["VarEps", "Mode"]), 2)`% of the variation is explained by the covariates. We are both in broad agreement.

@WhiteMEE2023 find that precipitation (`gmap`) is the most important variable, whereas we find temperature to be more important in both analyses. This is less of a conflict than it appears, as they state

    Holding other covariates constant in the South Africa data, annual average 
    precipitation has almost no estimated impact on beta diversity until the 
    average exceeds 300 mm according to estimates from the spGDMM. Beyond 300 mm, 
    annual precipitation has a very strong impact on beta diversity. This suggests 
    that precipitation averages below 300 mm are effectively similar in terms of 
    beta diversity. 

In the data only `r round(100*mean(SAData$gmap>300), 1)`% of the locations have a rainfall over 300 mm, which suggests their results are being driven by a small part of the environmental space. This also suggests that non-linear terms might be worth exploring, although the variation may also be driven by linear effects of the species that prefer higher rainfall.

### Example 2: The Inevitable BCI Data

This is counts of trees on an island that only exists to let ecologists take a holiday in the jungle. The data can be downloaded from @Condit2002.

```{r, GetDataBCI}

BCI.Env <- read.csv("data/Panama_env.csv")
BCI.Sp <- read.csv("data/Panama_species.csv")

# Stupid naming 
# "species" is actually site. And the format of the names is different between files
# This is why "data scientist" has a lower job satisfaction than "statistician"
BCI.Sp$site.no. <- sprintf("P%02d", as.numeric(gsub("P", "", BCI.Sp$species)))
BCI.Sp$species <- NULL

BCI.EnvCov <- c("precip", "elev", "age")
Pres <- apply(BCI.Sp, 2, function(x) sum(x>0))
sp.tmp <- BCI.Sp[,which(Pres>10)]
dat.tmp <- cbind(BCI.Env, sp.tmp)
SpNames <- names(sp.tmp)[!names(sp.tmp)%in%c("species", "site.no.")]
BCIData <- reshape(dat.tmp, direction = "long", varying = SpNames, v.names = "Count")
BCIData$Species <- factor(SpNames[BCIData$time])
BCIData$Site <- factor(BCIData$site.no.)
BCIData$SiteSpecies <- paste0(BCIData$Site, BCIData$Species)
# Scale env
Scale.tmp <- apply(BCIData[,BCI.EnvCov], 2, scale)
colnames(Scale.tmp) <- paste0(colnames(Scale.tmp), ".sc")
BCIData <- cbind(BCIData, Scale.tmp)
```

We can fit the model in INLA:

```{r, FiModelBCI}
#| warning: false


BCI.Formula <- formula(
  paste0("Count ~ f(Species, hyper = PCHyper) + f(Site, hyper = PCHyper) + ", 
         paste(
           c(paste0(BCI.EnvCov, ".sc"), 
             sapply(paste0(BCI.EnvCov, ".sc"), function(eu) {
               paste0("f(Species.", eu, ", ", eu, ", hyper = PCHyper)")
             })), collapse=" + "), " + f(OD, hyper = PCHyper)")
)

BCIenvInd <- data.frame(sapply(BCI.EnvCov, function(nm, dat) assign(nm, dat), 
                               dat=BCIData$Species))
names(BCIenvInd) <- paste0("Species.", names(BCIenvInd), ".sc")

BCIData <- cbind(BCIData, BCIenvInd)
BCIData$OD <- factor(1:nrow(BCIData))

# Linear combinations
# Should improve code
LCs.lst <- lapply(paste0(BCI.EnvCov, ".sc"), CreateLinComb, Levels=BCIData$Species)
BCI.LCs <- do.call(c, LCs.lst)

# Fit the model
BCIinla <- inla(BCI.Formula, data=BCIData, control.compute=list(config = TRUE),
                family="poisson", lincomb = BCI.LCs)
BCISumm <-summary(BCIinla)

```

```{r, CalcStatsBCI}
#| eval: true
#| fig-height: 3

BCIsamps <- inla.posterior.sample(nsim, BCIinla)

BCI.Posts <- data.frame(t(sapply(BCIsamps, CalcPostVar, data=BCIData)))
Use <- !names(BCI.Posts)%in%paste0("Precision.for.", c("Species", "Site"))
BCI.PropVar <- sweep(BCI.Posts[,Use], 1, rowSums(BCI.Posts[,Use]), "/")
names(BCI.PropVar)[3] <- "VarEps"

par(mfrow=c(1,3))
plot(density(BCI.PropVar$MeanDelta), main="Mean Covariate")
plot(density(BCI.PropVar$VarDelta), main="Variation in Covariate")
plot(density(BCI.PropVar$VarEps), main="Residual")

BCI.SummVars <- cbind(
  Mode=MCMCglmm::posterior.mode(coda::as.mcmc(BCI.PropVar)),
  coda::HPDinterval(coda::as.mcmc(BCI.PropVar), prob = 0.95)
)

```

We find that `r round(100*(1-BCI.SummVars["VarEps", "Mode"]), 2)`% of the variation in beta diversity is explained by the covariates, with a 95% HPDI of `r round(100*(1-BCI.SummVars["VarEps", "upper"]), 2)`% - `r round(100*(1-BCI.SummVars["VarEps", "lower"]), 2)`%.


```{r PlotEffectsBCI}

# Plot means
Use.fixed <- which(!grepl("\\(Int", rownames(BCISumm$fixed)))
Use.hyper <- sapply(BCI.EnvCov, function(x, y) grep(x, y),
                    y=names(BCIinla$marginals.hyperpar))

BCIHyperSDs <- CalcHyperSDs(BCIinla)

par(mar=c(4.1,1,3,1), oma=c(1,8,1,1), mfrow=c(1,2))
 PlotFixed(fixed=BCISumm$fixed, use=Use.fixed)
 PlotHyper(SDs=BCIHyperSDs, use=Use.hyper)

```

We can see that elevation has the largest average effect, with a negtive gradient, but there is variation across species in all effects.


```{r PlotCovarEffsBCI}
#| fig-height: 8
#| results: hide

BCI.RanEff <- BCIinla$summary.lincomb.derived

UseSumRan <- grep("Species\\.", names(BCIinla$summary.random))
Xlims <- range(BCI.RanEff[,c("0.025quant", "0.975quant")])

par(mar=c(2.1,1,3,1), mfrow=c(1,3))
sapply(BCI.EnvCov, function(env, RanEff, Fixed, ...) {
  PlotEff(summ=RanEff[grep(env, rownames(RanEff)),], 
        mean=Fixed[grep(env, rownames(Fixed)),],
        ylab="", xlab="", main=env, ...)
}, RanEff=BCI.RanEff, Fixed=BCISumm$fixed, xlim=Xlims)

```

Here we can see that almost all species are negatively affectsed by gradient, whereas the effects of precipitation and age as more mixed.

### Example 3: Australian Plants

This is occurrence of plants in Southwest Australia, the `southwest` data in the `gdm` package. The data are not well documented, with several presences per species per site. Absence data is not given. Fortunately, the environmental data does not vary within a site.

```{r, GetDataSW}
data("southwest")

SW.EnvCov <- names(southwest)[!names(southwest)%in%c("species","site", "Lat", "Long")]

Pres <- table(southwest$species)
sp.keep <- names(Pres[Pres>50]) #reduce to 10?

southwest.small <- southwest[southwest$species%in%sp.keep,]
southwest.small$species <- factor(southwest.small$species)
SW.pres <- expand.grid(species=unique(southwest.small$species),
                       site=unique(southwest.small$site))
SW.pres$Presence <- apply(SW.pres, 1, function(wh, dat) {
 any(dat$site==wh["site"] & dat$species==wh["species"])
}, dat=southwest.small)

# Extract environmental variables
SWenv <- data.frame(apply(southwest.small[,SW.EnvCov], 2, function(x, dat) {
  tapply(x, list(dat$site), function(xx) xx[1])
}, dat=southwest.small))
env.site <- rownames(SWenv)
SWenv <- data.frame(apply(SWenv, 2, scale))
names(SWenv) <- paste0(colnames(SWenv), ".sc")
SWenv$site <- env.site

SW.data <- merge(SW.pres, SWenv)
SW.data$Presence <- as.numeric(SW.data$Presence)

```

We can fit the model in INLA:

```{r, FiModelSW}
#| warning: false

SW.Formula <- formula(
  paste0("Presence ~ f(species, hyper = PCHyper) + f(site, hyper = PCHyper) + ", 
         paste(
           c(paste0(SW.EnvCov, ".sc"), 
             sapply(paste0(SW.EnvCov, ".sc"), function(eu) {
               paste0("f(Species.", eu, ", ", eu, ", hyper = PCHyper)")
             })), collapse=" + "))
)

SWenvInd <- data.frame(sapply(SW.EnvCov, function(nm, dat) assign(nm, dat), 
                               dat=SW.data$species))
names(SWenvInd) <- paste0("Species.", names(SWenvInd), ".sc")
SW.data <- cbind(SW.data, SWenvInd)

# Create linear combinations
LCs.lst <- lapply(paste0(SW.EnvCov, ".sc"), CreateLinComb, Levels=SW.data$species)
SW.LCs <- do.call(c, LCs.lst)

# Fit the model
SWinla <- inla(SW.Formula, data=SW.data, control.compute=list(config = TRUE),
                family="binomial", Ntrials = 1, lincomb = SW.LCs)
SWSumm <-summary(SWinla)
```

And now...

```{r, CalcStatsSW}
#| eval: true

SWsamps <- inla.posterior.sample(nsim, SWinla)

SW.Posts <- data.frame(t(sapply(SWsamps, CalcPostVar, data=SW.data)))
SW.Posts$logitVar <- pi/3
Use <- !names(SW.Posts)%in%paste0("Precision.for.", c("species", "site"))
SW.PropVar <- sweep(SW.Posts[,Use], 1, rowSums(SW.Posts[,Use]), "/")
names(SW.PropVar)[3] <- "VarEps"

par(mfrow=c(1,3))
plot(density(SW.PropVar$MeanDelta), main="Mean Covariate")
plot(density(SW.PropVar$VarDelta), main="Variation in Covariate")
plot(density(SW.PropVar$VarEps), main="Residual")

SW.SummVars <- cbind(
  Mode=MCMCglmm::posterior.mode(coda::as.mcmc(SW.PropVar)),
  coda::HPDinterval(coda::as.mcmc(SW.PropVar), prob = 0.95)
)

```

We find that `r round(100*(1-SW.SummVars["VarEps", "Mode"]), 2)`% of the variation in beta diversity is explained by the covariates, with a 95% HPDI of `r round(100*(1-SW.SummVars["VarEps", "upper"]), 2)`% - `r round(100*(1-SW.SummVars["VarEps", "lower"]), 2)`%.

```{r PlotEffectsSW}

# Plot means
Use.fixed <- which(!grepl("\\(Int", rownames(SWSumm$fixed)))
Use.hyper <- sapply(SW.EnvCov, function(x, y) grep(x, y),
                    y=names(SWinla$marginals.hyperpar))

SWHyperSDs <- CalcHyperSDs(SWinla)

par(mar=c(4.1,1,3,1), oma=c(1,8,1,1), mfrow=c(1,2))
 PlotFixed(fixed=SWSumm$fixed, use=Use.fixed)
 PlotHyper(SDs=SWHyperSDs, use=Use.hyper)

```

We can see that elevation has the largest average effect, but there is variation across species in all effects


```{r PlotCovarEffsSW}
#| fig-height: 20
#| results: hide

SW.RanEff <- SWinla$summary.lincomb.derived

UseSumRan <- grep("Species\\.", names(SWinla$summary.random))
Xlims <- range(SW.RanEff[,c("0.025quant", "0.975quant")])

par(mar=c(2.1,1,3,1), mfrow=c(5,2))

sapply(SW.EnvCov, function(env, RanEff, Fixed, ...) {
  PlotEff(summ=RanEff[grep(env, rownames(RanEff)),], 
        mean=Fixed[grep(env, rownames(Fixed)),],
        ylab="", xlab="", main=env, ...)
}, RanEff=SW.RanEff, Fixed=SWSumm$fixed, xlim=Xlims)

```


## References
