---
title: "German Butterfly Restoration"
author: "Bob O'Hara"
format: html
---

## Quarto

Data from Zenodo: https://zenodo.org/records/20763922
Pangaea: https://doi.pangaea.de/10.1594/PANGAEA.982396
Paper: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2664.70513





::: {.callout-note collapse=true}

## Observation Model

The data are counts of each species, and we model both presence/absence and counts. We have multiple replicate observations per site, so we assume the counts, $n_{kj}$ on replicate $k$ for species $j$, follow a zero inflated Poisson lognormal distribution with mean $\lambda_{ij}$ (i.e. constant across replicates within a site). We have a Gaussian overdispersion term for excess variation within a replicate that represents sampling variation. For presences (i.e. to account for zero inflation) we use an occupancy model. 
## Downloading and formatting the data

The data are abundances (counts) on each location on a transect. There are 3 islands, each with 3 transects, each with 3 stations. Each station was sampled in 2 seasons, and has 3 replicates.

The steps are:
1. import the data

```{r}
#| label: ImportData
#| echo: false
butterfly_data_raw <- read.csv("data/data_butterfly_raw.csv")

# Remove April and "additionally found species" (not clear what these are.
# Note that the original code does not remove them, as it removes "zuzatz")
butterfly_data <- butterfly_data_raw[butterfly_data_raw$observation != "April" &
                                       butterfly_data_raw$sub.transect != "Zusatz",]
# change to make code easier to follow
names(butterfly_data)[names(butterfly_data)=="observation"] <- "month"

# change id.sub.transect so it is easier to parse with regular expressions
butterfly_data$id.sub.transect <- paste(butterfly_data$id.site, butterfly_data$sub.transect, sep=".")

# For some reason there are multiple observations of species on the same transect/month. So add these up.
# tabl <- table(butterfly_data$butterfly.name, butterfly_data$sub.transect)
butterfly_data.agg <- aggregate(butterfly_data$butterfly.no, 
                                by=list(id.sub.transect=butterfly_data$id.sub.transect, 
                                        butterfly.name=butterfly_data$butterfly.name), FUN=sum)

# Change from wide to long format
Counts <- reshape(data=butterfly_data.agg, v.names = "x",
                            idvar = "id.sub.transect", timevar = "butterfly.name", 
                            direction="wide")
Counts[is.na(Counts)] <- 0
names(Counts) <- gsub("x.", "", names(Counts), fixed = TRUE)

Counts$Site <- as.factor(gsub("\\..*", "", Counts$id.sub.transect))
# Get species names
Species <- unique(butterfly_data$butterfly.name)
# Check to make sure these are also comumn names
if(mean(Species%in%names(Counts))<1) message("Some species not in Count")

# Get presence fromdata
Present <- apply(Counts[,Species], 2, function(Spp, Site) 
  tapply(Spp, list(Site), function(a) c(NA,1)[1+any(a>0)]), 
                 Site = Counts$Site)

```

2. Format the data for JAGS.

```{r}
#| label: FormatForJAGS
#| echo: false
DataToJAGS <- list(
  Abund = as.matrix(Counts[,Species]), 
  Site = as.numeric(Counts$Site),
  Present = Present, 
#  temperature = as.vector(scale(tapply(Env$temperature, list(Counts$Site), mean))),
  # Island <- Counts$Island, # Might use these later
  # Transect <- Counts$Transect, 
  # Station <- Counts$Station, 
  # Season <- Counts$Season

  NSp = length(Species),
  NSite = nlevels(Counts$Site),
  NRep = nrow(Counts)
)

# Function to simulate initial values
GetInits <- function(data, corr=FALSE) {
  res <- list(
    mu.be.lam = rnorm(1, 0, 0.1),
    mu.be.z = rnorm(1, 0, 0.1),
    mu.al.lam = rnorm(1, 0, 0.1),
    mu.al.z = rnorm(1, 0, 0.1),
    mu.lam = rnorm(1, 0, 0.1),
    mu.z = rnorm(1, 0, 0.1),
    
    sd.be.lam = rgamma(1, 100, 500),
    sd.be.z = rgamma(1, 100, 500),
    sd.al.lam = rgamma(1, 100, 500),
    sdOD = rgamma(1, 100, 500),
    sd.al.z = rgamma(1, 100, 500))
  if(exists("temperature", data)) {
    res$mu.gam.lam = rnorm(1, 0, 1)
    res$sd.gam.lam = rgamma(1, 100, 500)
    res$mu.gam.z = rnorm(1, 0, 0.5)
    res$sd.gam.z = rgamma(1, 100, 500)
  }
  if(corr) {
    res$rho = rnorm(data$NSp, 0, 0.1)
    res$VarSiteOD = matrix(rgamma(2*data$NSp, 100, 100), ncol=2)
  }
  res
}

```

3. Write the model in JAGS code

```{r}
#| label: JAGScode
#| echo: false
cat("
model {
  for(sp in 1:NSp) {
    for(rep in 1:NRep) {
      Abund[rep, sp] ~ dpois(lambda[rep, sp])
      lambda[rep, sp] <- mu[rep, sp]*Present[Site[rep], sp]
      OD[rep, sp] ~ dnorm(0, tauOD)
      log(mu[rep, sp]) <- logmu[Site[rep], sp] + OD[rep, sp]
    }
    for(site in 1:NSite) {
      Present[site, sp] ~ dbern(p[site, sp])
      SiteOD[site, sp] ~ dmnorm(0, tauSiteOD)
      logmu[site, sp] <- mu.lam + al.lamST[site] + be.lamST[sp] + SiteOD[site, sp]
      logit(p[site, sp]) <- mu.z + al.zST[site] + be.zST[sp]
    }
    be.zST[sp] ~ dnorm(mu.be.z, tau.be.z)
    be.lamST[sp] ~ dnorm(mu.be.lam, tau.be.lam)

    be.lam[sp] <- be.lamST[sp] - mean(be.lamST[])
    be.z[sp] <- be.zST[sp] - mean(be.zST[])
  }
  for(site in 1:NSite) {
    al.lamST[site] ~ dnorm(mu.al.lam, tau.al.lam)
    al.zST[site] ~ dnorm(mu.al.z, tau.al.z)
    al.lam[site] <- al.lamST[site] - mean(al.lamST[])
    al.z[site] <- al.zST[site] - mean(al.zST[])
    SR[site] <- sum(Present[site, ])
  }
    
  mu.be.lam ~ dnorm(0, 1)
  mu.be.z ~ dnorm(0, 1)
  mu.al.lam ~ dnorm(0, 1)
  mu.al.z ~ dnorm(0, 1)
  mu.lam ~ dnorm(0, 1)
  mu.z ~ dnorm(0, 1)

  Mean.lam <- mu.lam + mean(al.lamST[]) + mean(be.lamST[])
  Mean.z <- mu.z + mean(al.zST[]) + mean(be.zST[])

  tau.be.lam <- pow(sd.be.lam, -2); sd.be.lam ~ dexp(1)
  tau.be.z <- pow(sd.be.z, -2); sd.be.z ~ dexp(1)
  tau.al.lam <- pow(sd.al.lam, -2); sd.al.lam ~ dexp(1)
  tau.al.z <- pow(sd.al.z, -2); sd.al.z ~ dexp(1)

  tauOD <- pow(sdOD, -2); sdOD ~ dexp(1)
  tauSiteOD <- pow(sdSiteOD, -2); sdSiteOD ~ dexp(1)

# Finite population variances
  StdDev.be.lam <- sd(be.lamST[])
  StdDev.be.z <- sd(be.zST[])
  StdDev.al.lam  <- sd(al.lamST[])
  StdDev.al.z <- sd(al.zST[])

  StdDevSiteOD <- sd(SiteOD[, ])
  StdDevDelta.lam <- sd(SiteOD[, ]) # interaction
  StdDevDelta.z <- sqrt(3.289868) # interaction
  StdDevOD <- sd(OD[, ]) # observation level OD
  StdDevTotal.z <- sqrt(StdDev.al.z^2 + StdDev.be.z^2 + StdDevDelta.z^2)
  StdDevTotal.lam <- sqrt(StdDev.al.lam^2 + StdDev.be.lam^2 + StdDevDelta.lam^2)

}", file="BeatingBetaCode.jag")

```



4. Run the JAGS code

```{r}
#| label: FitModel
#| warning: false
#| message: false
#| output: false
#| echo: false
library(runjags)
library(coda)

ToMonitor.z <- c("Mean.z", "SR", "sd.al.z", "sd.be.z", "al.z")
ToMonitor.lam <- c("Mean.lam", "sd.al.lam", "sd.be.lam", "sdSiteOD", "sdOD", "StdDevDelta")
ToMonitor.gam <- c("mu.gam.lam", "sd.gam.lam", "sd.gam.z", "mu.gam.z", 
                   "gamma.lam", "gamma.z", "Mean.gam.lam", "Mean.gam.z")
ToMonitor.means <- c("logmu", "Present", "p") # if we want everything
ToMonitor.DivStats <- c("Unevenness", "UnevennessOcc")
ToMonitor.Unevenness <- c("be.lam", "SiteOD", "Present")

ToMonitor <- unique(c(ToMonitor.z, ToMonitor.lam, ToMonitor.gam, ToMonitor.means,
                      ToMonitor.DivStats))
# Add to this as we go on
ToMonitor <- unique(c(ToMonitor.means))

Inits <- replicate(2, GetInits(DataToJAGS), simplify = FALSE)

DivModel <- run.jags(model = "BeatingBetaCode.jag", monitor=ToMonitor, data=DataToJAGS,
                  inits = Inits, method="parallel", silent.jags=TRUE,
#         n.chains = 2, adapt = 1e2, burnin = 0, sample = 150)
#         n.chains = 4, adapt = 1e3, burnin = 0, sample = 5e3)
         n.chains = 6, adapt = 1e3, burnin = 0, sample = 1e3)
# Summ <- summary(DivModel, vars= c(ToMonitor.z, ToMonitor.lam, ToMonitor.gam))
# Summ <- summary(DivModel, vars= ToMonitor)

# Combine MCMC into one data fame, so it's easier to work with below
DivPars <- combine.mcmc(DivModel)



```

We get some output, which we will use to estimate our diversity statistics

:::






# Diversity Output

All of our diversity metrics are based on decomposing the distributions into:

- Occurrence
- Abundance

And looking at the means and variances of the parameters associated with these.

```{r}
#| label: DiversityFunctions
#| code-fold: true
#| code-summary: To calculate the statiastics need some functions, which are hidden away here

Sites <- 1:DataToJAGS$NSite
Species <- 1:DataToJAGS$NSp

# Function to convert vector of data to matrix

# vec: vector of values, should have names. 
# varname: name of variable to extract
# RowsFirstIndex: Should the first index be for rows or columns?
# sortNumeric: Should the indices be converted to integer and then sorted? Warning: make sure they are integers!

# Details
# For the variable of interest the name should be of the format "varname[i,j]" 
#  where i and j are the row and column names (in either order). 
#  They will probably be integers.
VecToMatrix <- function(vec, varname, RowsFirstIndex=TRUE, sortNumeric=TRUE) {
  # extract variable and indices
  variable <- vec[grep(paste0("^", varname, "\\["), names(vec))]
  Index1 <- gsub(",.*", "", gsub(".*\\[", "", names(variable)))
  Index2 <- gsub("\\]", "", gsub(".*,", "", names(variable)))

# create matrix with row and column names 
  mat <- matrix(NA, nrow=length(unique(Index1)), ncol=length(unique(Index2)),
                dimnames=list(unique(Index1), unique(Index2)))
# fill in entries
  for(i in seq_along(variable))  mat[Index1[i], Index2[i]] <- variable[i]
# If the column name is the first index, transpose the matrix
  if(!RowsFirstIndex) mat <- t(mat)
# Sort the rows and columns, if requested
  if(sortNumeric) {
    mat2 <- mat[sort(as.numeric(rownames(mat))), 
                sort(as.numeric(colnames(mat)))]
  }
  return(mat)
}


MatrixToArray <- function(mat, varname, RowsSecondIndex=TRUE, sortNumeric=TRUE) {
  # extract variable and indices
  variable <- mat[,grep(paste0("^", varname, "\\["), colnames(mat))]
  Index1 <- gsub(",.*", "", gsub(".*\\[", "", colnames(variable)))
  Index2 <- gsub("\\]", "", gsub(".*,", "", colnames(variable)))

# create matrix with row and column names 
  arr <- array(NA, 
               dim =c(nrow(mat), length(unique(Index1)), length(unique(Index2))),
               dimnames=list(rownames(mat), unique(Index1), unique(Index2)))
# fill in entries
  for(i in seq_along(variable[1,]))  arr[,Index1[i], Index2[i]] <- variable[,i]
# If the column name is the first index, transpose the matrix
  if(!RowsSecondIndex) arr <- aperm(1,3,2)
# Sort the rows and columns, if requested
  if(sortNumeric) {
    arr <- arr[,sort(as.numeric(dimnames(arr)[[2]])), 
                sort(as.numeric(dimnames(arr)[[3]]))]
  }
  return(arr)
}



# this is a function to calculate the summaries we want
CalcSumm <- function(mc, ...) {
  require(coda)
  require(MCMCglmm)
  Summ <- summary(mc)
  Summ$HPD <- HPDinterval(mc, prob=0.95)
  Summ$PMode <- posterior.mode(mc, ...)
  Summ
}

# Calculate summary stats from array: keep as mcmc object
CalcSummaryFromArray <- function(arr, FUN=mean, mcmc=FALSE, centre=TRUE) {
  Grand <- apply(arr, "Iteration", FUN, na.rm=TRUE)
  Sp <- apply(arr, c("Iteration", "Species"), FUN, na.rm=TRUE)
  colnames(Sp) <- paste0("Species", colnames(Sp))
  Site <- apply(arr, c("Iteration", "Site"), FUN, na.rm=TRUE)
  colnames(Site) <- paste0("Site", colnames(Site))
  if(centre) {
    Sp <- sweep(Sp, 1, Grand, "-")
    Site <- sweep(Site, 1, Grand, "-")
  }
  res <- cbind(Grand, Sp, Site)
  if(mcmc) res <- coda::as.mcmc(res)
  res
}

# Calculate correlation between statistics in two data sets 
#   (err, that could be explained better)
CalcCorr <- function(mcmc1, mcmc2, Name) {
  if(nrow(mcmc1)!=nrow(mcmc2)) stop("mcmc1 & mcmc2 must have same number of rows")
  Use1 <- colnames(mcmc1)[grep(Name, colnames(mcmc1))]
  Use2 <- colnames(mcmc2)[grep(Name, colnames(mcmc2))]
  if(any(Use1%notin%Use2) | any(Use2%notin%Use1)) stop(paste0("mcmc1 and mcmc2 should have same variables for ", Name))
  
  Cor <-   sapply(1:nrow(mcmc1), function(i, mc1, mc2) {
    cor(mc1[i,], mc2[i,])
  }, mc1 = mcmc1[,Use1], mc2 = mcmc2[,Use2])  
  Cor
}


# Function to plot parameters
PlotPars <- function(Summ, At="Orig", use=NULL, ave=NULL, Xlim=NULL, ...) {
  if(is.null(use)) use <- 1:length(Summ$PMode)
  if(length(At)==length(use)) {
    At.y <- At
  } else {
    At.y <- switch(At, 
                   Orig = 1:use, 
                   rank = rank(Summ$PMode[use], ties="random")
    )
  }
  if(is.null(Xlim)) Xlim=range(Summ$HPD[use,])
  plot(Summ$PMode[use], At.y, xlim=Xlim, type="n", ...)
  if(!is.null(ave)) {
  rect(Summ$HPD[ave,"lower"], -5, Summ$HPD[ave,"upper"], 2*max(At.y), border = NA, col="pink")
  abline(v=Summ$PMode[ave], col=2)
    
  }
  points(Summ$PMode[use], At.y, ...)
  segments(Summ$HPD[use,"lower"], At.y, Summ$HPD[use,"upper"], At.y)
  
}


```

## Diversity Means

## $S_i$, $\alpha$-mean diversity: Species richness and Mean Abundance

For occupancy this relates to species richness. Because our fitted model is on the logit scale, we can calculate $S_i$ for occupancy on that scale too, as well as species richness itself.

For abundance we only average over the species that are present: for those we do not observe, we estimate presence.

```{r}
#| label: CalcAlphaMeanDiv
#| code-fold: true

p.arr <- MatrixToArray(DivPars, varname = "p")
lgtp.arr <- log(p.arr/(1 - p.arr))
Present.arr <- MatrixToArray(DivPars, varname = "Present")
abund.arr <- MatrixToArray(DivPars, varname = "logmu")
names(dimnames(lgtp.arr)) <- names(dimnames(Present.arr)) <- 
  names(dimnames(abund.arr)) <- c("Iteration", "Site", "Species")
abund.arr[Present.arr==0] <- NA


lgtp.means <- CalcSummaryFromArray(lgtp.arr, mcmc=TRUE)
Present.sum <- CalcSummaryFromArray(Present.arr, FUN=sum, mcmc=TRUE, centre = FALSE)
abund.means <- CalcSummaryFromArray(abund.arr, mcmc=TRUE)

lgtp.means.summ <- CalcSumm(lgtp.means)
Present.sum.summ <- CalcSumm(Present.sum)
abund.means.summ <- CalcSumm(abund.means)

MeansCorrelations <- as.mcmc(cbind(
  lgtp.abund.sp.cor =  CalcCorr(mcmc1 = lgtp.means, mcmc2 = abund.means, Name = "Species"),
  lgtp.Present.sp.cor =  CalcCorr(mcmc1 = lgtp.means, mcmc2 = Present.sum, Name = "Species"),
  Present.abund.sp.cor =  CalcCorr(mcmc1 = Present.sum, mcmc2 = abund.means, Name = "Species"),
  
  lgtp.abund.site.cor =  CalcCorr(mcmc1 = lgtp.means, mcmc2 = abund.means, Name = "Site"),
  lgtp.Present.site.cor =  CalcCorr(mcmc1 = lgtp.means, mcmc2 = Present.sum, Name = "Site"),
  Present.abund.site.cor =  CalcCorr(mcmc1 = Present.sum, mcmc2 = abund.means, Name = "Site")
))
MeansCorrelations.summ <- CalcSumm(MeansCorrelations)

```

The estimates are plotted below, with the same site on the same row.

```{r}
#| label: PlotAlphaMeanDiv
#| code-fold: true
#| fig-height: 8
#| fig-cap: Posterior modes and 95% HPDIs for unevenness of communities on sites

At.mean <- rank(Present.sum.summ$PMode[grep("^Site", names(Present.sum.summ$PMode))], ties="random")
par(mfrow=c(1,3), mar=c(2,2,4,1))
PlotPars(lgtp.means.summ, use=grep("^Site", names(lgtp.means.summ$PMode)), At=At.mean, main="Mean Occupancy")
PlotPars(Present.sum.summ, use=grep("^Site", names(Present.sum.summ$PMode)), At=At.mean, main="Species Richness", 
         Xlim=c(0,DataToJAGS$NSp))
PlotPars(abund.means.summ, use=grep("^Site", names(abund.means.summ$PMode)), At=At.mean, main="Mean Abundance")

```

Not surprisingly, the posterior modes for species richness and mean occupancy are strongly correlated (posterior mode: 
`r round(MeansCorrelations.summ$PMode["lgtp.Present.sp.cor"], 2)`, 95% CI: 
`r round(MeansCorrelations.summ$quantiles["lgtp.Present.sp.cor","2.5%"], 2)` - 
`r round(MeansCorrelations.summ$quantiles["lgtp.Present.sp.cor","97.5%"], 2)`). 
And occurrence and abundance have a weak uncorrelation (posterior mode: 
`r round(MeansCorrelations.summ$PMode["lgtp.abund.sp.cor"], 2)`, 95% CI: 
`r round(MeansCorrelations.summ$quantiles["lgtp.abund.sp.cor","2.5%"], 2)` - 
`r round(MeansCorrelations.summ$quantiles["lgtp.abund.sp.cor","97.5%"], 2)`).

## $O_j$, $\beta$-mean diversity: Variation across sites for each species

We can also look at species. We can see that most species are rare (an observation that will not get us in to *Nature*), but a few are more common.


```{r}
#| label: PlotBetaMeanDiv
#| code-fold: true

At.sp <- rank(Present.sum.summ$PMode[grep("Species", names(Present.sum.summ$PMode))], ties="random")
par(mfrow=c(1,3), mar=c(2,2,4,1))
PlotPars(lgtp.means.summ, use=grep("Species", names(lgtp.means.summ$PMode)), At=At.sp, main="Mean Occupancy") #, Xlim=c(-50,200))
PlotPars(Present.sum.summ, use=grep("Species", names(Present.sum.summ$PMode)), At=At.sp, main="(Local) Range Size",
         Xlim=c(0,DataToJAGS$NSite), xaxs="i")
PlotPars(abund.means.summ, use=grep("Species", names(abund.means.summ$PMode)), At=At.sp, main="Mean Abundance")
```

For about a third of the species the estimated range size is low, but has a high uncertainty: these species also have low estimated abundance, so it's more likely to get a false negative.

The posterior modes for range size and mean occupancy are strongly correlated (posterior mode: 
`r round(MeansCorrelations.summ$PMode["lgtp.Present.site.cor"], 2)`, 95% CI: 
`r round(MeansCorrelations.summ$quantiles["lgtp.Present.site.cor","2.5%"], 2)` - 
`r round(MeansCorrelations.summ$quantiles["lgtp.Present.site.cor","97.5%"], 2)`). 
But the occurrence and abundance are uncorrelated (posterior mode: 
`r round(MeansCorrelations.summ$PMode["lgtp.abund.site.cor"], 2)`, 95% CI: 
`r round(MeansCorrelations.summ$quantiles["lgtp.abund.site.cor","2.5%"], 2)` - 
`r round(MeansCorrelations.summ$quantiles["lgtp.abund.site.cor","97.5%"], 2)`).


## Diversity Variances

```{r}
#| label: CalculateVariances
#| code-fold: true
lgtp.var <- CalcSummaryFromArray(lgtp.arr, FUN=sd, mcmc=TRUE, centre = FALSE)^2
Present.var <- CalcSummaryFromArray(Present.arr, FUN=sd, mcmc=TRUE, centre = FALSE)^2
abund.var <- CalcSummaryFromArray(abund.arr, FUN=sd, mcmc=TRUE, centre = FALSE)^2
abund.var[is.na(abund.var)] <- 0

# Add on mean variances
lgtp.var <- as.mcmc(cbind(lgtp.var, 
                          MeanSpecies = apply(lgtp.var[,grep("^Species", colnames(lgtp.var))], 1, mean),
                          MeanSite = apply(lgtp.var[,grep("^Site", colnames(lgtp.var))], 1, mean)))
Present.var <- as.mcmc(cbind(Present.var, 
                             MeanSpecies = apply(Present.var[,grep("^Species", colnames(Present.var))], 1, mean),
                             MeanSite = apply(Present.var[,grep("^Site", colnames(Present.var))], 1, mean)))
abund.var <- as.mcmc(cbind(abund.var, 
                           MeanSpecies = apply(abund.var[,grep("^Species", colnames(abund.var))], 1, mean),
                           MeanSite = apply(abund.var[,grep("^Site", colnames(abund.var))], 1, mean)))

# lgtp.var.summ is horribly right skewed, so we need to change the defaults for the smoothing used to estimate the posterior mode
lgtp.var.summ <- CalcSumm(lgtp.var, from=0, to=1000, bw=0.1)
Present.var.summ <- CalcSumm(Present.var)
abund.var.summ <- CalcSumm(abund.var)

```


### $\alpha$-var-diversity: Unevenness

Unevenness for each site is $U_j = \text{Var}_i(k_{i,j})$, i.e. the variance in the species' effects on each site. For abundance this is related to measures such as Shannon's and Simpson's diversity: the more evenly distributed species' abundances are, the higher the diversity (an uneven distribution would mean one or a few species with high abundances, i.e. dominating the community). Our variance is a measure of unevenness: when it is zero, all species have the same abundance, so there is maximal diversity.

```{r}
#| label: PlotUnevenness
#| code-fold: true
#| warning: false

# Calculate mean unevenness
At.site <- rank(Present.var.summ$PMode[grep("^Site", names(Present.var.summ$PMode))], ties="random")
par(mfrow=c(1,3), mar=c(2,2,4,1), oma=c(2,2,0,0))
PlotPars(lgtp.var.summ, use=grep("^Site", names(lgtp.var.summ$PMode)), 
         At=At.site, main="Mean Occupancy", ave="MeanSite") #log="x", 
PlotPars(Present.var.summ, use=grep("^Site", names(Present.var.summ$PMode)), 
         At=At.site, main="(Local) Range Size", ave="MeanSite")
PlotPars(abund.var.summ, use=grep("^Site", names(abund.var.summ$PMode)), 
         At=At.site, main="Unevenness in Abundance", ave="MeanSite")
mtext("Variance", 1, outer=TRUE)
mtext("Site", 2, outer=TRUE)

```

We can see that unevenness has some variation in abundance. For mean occupancy the unevenness variance is the same for all sites, for some reason I need to check. The variation in local range soze variance is thus because of changes in the mean.

With a lot of rare species, occupancy is uncertain, so the unevenness is perhaps to be expected.

## $\sigma^2_\eta$, $\alpha$ Disparity, and its decomposition. What was pickiness.

This is the variance in $O_j$, the average prevalence of a species.

```{r}
#| label: PlotWhatWasPickiness
#| code-fold: true
#| warning: false

# Calculate mean unevenness
At.sp <- rank(Present.var.summ$PMode[grep("^Species", names(Present.var.summ$PMode))], ties="random")
par(mfrow=c(1,3), mar=c(2,2,4,1), oma=c(2,2,0,0))
PlotPars(lgtp.var.summ, use=grep("^Species", names(lgtp.var.summ$PMode)), 
         At=At.sp, main="Mean Occupancy", ave="MeanSpecies") #log="x", 
PlotPars(Present.var.summ, use=grep("^Species", names(Present.var.summ$PMode)), 
         At=At.sp, main="(Local) Range Size", ave="MeanSpecies")
PlotPars(abund.var.summ, use=grep("^Species", names(abund.var.summ$PMode)), 
         At=At.sp, main="Unevenness in Abundance", ave="MeanSpecies")
mtext("Variance", 1, outer=TRUE)
mtext("Site", 2, outer=TRUE)

```

For incidence, it looks a but of a mess. For abundance, the species only present at a few sites have poor estimates: the others seem roughly similar.


# Gamma Diversity and its decomposition

The statistics presented above are site or species specific. We can look at overall diversity by looking across sites, to get (reading from Erik's figure)

$$
\begin{aligned}
\sigma^2_\gamma &= \sigma^2_\alpha + \sigma^2_\delta + \sigma^2_\beta \\
 &= R + D \\
\sigma^2_\gamma &= \sigma^2_\beta + \sigma^2_\delta + \sigma^2_\alpha \\
 &= U + N 
\end{aligned}
$$

where 

- $R = \sigma^2_\alpha + \sigma^2_\delta = \bar{P}_j$ is replacement
- $D = \sigma^2_\beta = \text{Var}(O_{j})$ is disparity. Variance over species in mean prevalence etc.
- $U = \sigma^2_\beta + \sigma^2_\delta = \bar{U}_i$ is mean unevenness
- $N = \sigma^2_\alpha = \text{Var}(S_{i})$: is nestedness. Variance over sites in mean prevalence, species richness

```{r}
#| label: CalcGammaStats
#| code-fold: true
#| warning: false

# N <- apply(abund.arr, c("Iteration", "Site"), mean)
# N <- apply(apply(abund.arr, c("Iteration", "Site"), mean, na.rm=TRUE), "Iteration", sd)^2
# D <- apply(apply(abund.arr, c("Iteration", "Species"), mean, na.rm=TRUE), "Iteration", sd)^2
CalcGammaStats <- function(arr) {
  GammaStats <- cbind(U = apply(apply(arr, c("Iteration", "Site"), sd, na.rm=TRUE)^2, 
                                "Iteration", mean, na.rm=TRUE)^2,
                      R = apply(apply(arr, c("Iteration", "Species"), sd, na.rm=TRUE)^2, 
                                "Iteration", mean, na.rm=TRUE)^2,
                      N = apply(apply(arr, c("Iteration", "Site"), mean, na.rm=TRUE), 
                                "Iteration", sd, na.rm=TRUE)^2,
                      D = apply(apply(arr, c("Iteration", "Species"), mean, na.rm=TRUE), 
                                "Iteration", sd, na.rm=TRUE)^2)
  colnames(GammaStats) <- c("U", "R", "N", "D")
  GammaStats <- cbind(GammaStats, 
                      PropN = GammaStats[,"N"]/(GammaStats[,"U"]+GammaStats[,"N"]),
                      PropD = GammaStats[,"D"]/(GammaStats[,"D"]+GammaStats[,"R"]))
  GammaStats
  
}

Gamma.lgtp <- CalcGammaStats(lgtp.arr)
Gamma.pres <- CalcGammaStats(Present.arr)
Gamma.abund <- CalcGammaStats(abund.arr)

Summ.Gamma.lgtp <- CalcSumm(as.mcmc(Gamma.lgtp))
Summ.Gamma.pres <- CalcSumm(as.mcmc(Gamma.pres))
Summ.Gamma.abund <- CalcSumm(as.mcmc(Gamma.abund))

StatsForFigure <- data.frame(
  PMode = c(Summ.Gamma.lgtp$PMode[c("PropN", "PropD")], 
             Summ.Gamma.pres$PMode[c("PropN", "PropD")], 
             Summ.Gamma.abund$PMode[c("PropN", "PropD")]), 
  L95 = c(Summ.Gamma.lgtp$HPD[c("PropN", "PropD"),"lower"], 
           Summ.Gamma.pres$HPD[c("PropN", "PropD"),"lower"], 
             Summ.Gamma.abund$HPD[c("PropN", "PropD"),"lower"]), 
  U95 = c(Summ.Gamma.lgtp$HPD[c("PropN", "PropD"),"upper"], 
           Summ.Gamma.pres$HPD[c("PropN", "PropD"),"upper"], 
             Summ.Gamma.abund$HPD[c("PropN", "PropD"),"upper"]),
  At.y = rep(1:3, each=2),
  Col = rep(1:2, times=3)
)

```

From the plot we can see that nestedness is low, so most variation is across sites is because of turnover, rather than changes in total numbers of species or numbers of individuals. In contrast, disparity is high across species, especially abundance, so variation between species in global occupancy and abundance (i.e. some are rare and others common) is driving diversity in these communities, rather than species having similar overall abundances but segregating between sites.

```{r}
#| label: PlotGammaStats
#| code-fold: true
#| warning: false


par(mfrow=c(1,1), mar=c(4.1,6,1,1))
plot(StatsForFigure$PMode, StatsForFigure$At.y, xlim=c(0,1), yaxt="n", 
     col=StatsForFigure$Col, pch=16, xlab="Proportion", ylab="")
segments(StatsForFigure$L95, StatsForFigure$At.y, 
         StatsForFigure$U95, StatsForFigure$At.y, col=StatsForFigure$Col)
axis(2, c("Mean\nOccupancy", "Range\nSize", "Abundance"), las=1, at=1:3)
legend(0.21,2.7, c("Nestedness", "Disparity"), col=1:2, lty=1)

```

